---
title: "UPPMAX Account Guide"
subtitle: "Instructions to apply for an UPPMAX account and join the course project."
date: ""
author: ""
sidebar: false
code-tools: false
format: html
---

```{r}
#| include: false
library(yaml)
library(here)
upc <- yaml::read_yaml(here("_quarto.yml"))$uppmax_compute
```

:::{.callout-caution}
Do these steps well in advance as it can take up to 2 weeks for UPPMAX accounts to be approved. If this is incomplete, you may not be able to follow the labs during the workshop.
:::

These are the basic steps in this process:

1. Create an account in SUPR.
2.​ ​Apply for membership in the course project.
3.​ ​Accept the SNIC User Agreement in SUPR.
4.​ ​Apply for an UPPMAX account in SUPR.
5.​ ​Wait for an email from UPPMAX with your UPPMAX account details.
6.​ ​Log in with your new UPPMAX account.
7.​ ​Create a new file with your user name in the designated folder. This way we will
know who has completed these steps.

## ​Create an account in SUPR.

​If you already have a SUPR account, please continue to step 2.

​Go to​ https://supr.snic.se/​ and click “Register New Person” at the bottom of the first page. Complete the registration process, preferably using SWAMID, and login. If you for some reason can't use SWAMID to login, you will have to send physical (not electronic) copy of your ID to a place in Gothenburg for manual approval. Do this as ​soon as possible​, as this process can take ​more than 2 weeks.

## ​Apply for membership

- ​Log in using your SUPR account.
- ​Under the “Projects” heading, press the “View and Manage Projects” button.
- ​Search for the project ID: `r cat(upc)`
- ​Press the “Request” button to apply for membership in the project.

## ​Accept the SNIC User Agreement

- ​In SUPR, click on the link "Personal Information" in the left sidebar. You will have to accept the SNIC User Agreement to be able to get an UPPMAX account.

## Apply for an UPPMAX account

- ​In SUPR, click on the link "Accounts" in the left sidebar and apply for an UPPMAX account under the heading "Account Requests".

## UPPMAX account details

- ​Within about 2 working days you should get an email with instructions. ​Please, follow these instructions carefully.
- ​A while later you will get an email with your user name, and another email with a link to your password. 

:::{.callout-caution}
The link is only valid for ​1​ ​VISIT or 7 days​, so if you click the link you better save the password, because you will not be able to use the link again. Do this before 7 days have passed, otherwise the link will no longer be valid.
:::

## Log in with your new UPPMAX account

- ​Open your terminal program (Terminal in OSX and Linux, otherwise download MobaXterm​ (portable edition) if you have Windows).
- ​Type this command in your terminal program: `ssh username@rackham.uppmax.uu.se`
- ​You will be asked for your password now, and you will not see any response in the terminal while typing your password. This is to hide the length of your password, i.e. normal. Just press enter when you have typed it in and you should log in.
​If it is the first time you log in, it will ask you to change your LDAP password (the password you just typed). It will directly ask you for your password again, so type it once more. After that it will ask you for your new password, so make up a new one and press enter. After that it will ask you to confirm the new password. When the password change is completed you will be disconnected and you will have to connect again, using your new password to log in this time.

## ​Create a new file with your username in the designated folder.

- ​After having received information that your membership is approved, ​wait 24 h before continuing, as it takes up to 24 h for SUPR to sync with UPPMAX. Else, you will get the message “permission denied” if you try to create the file before this sync has been performed.

```{r}
#| attr-output: "filename='bash'"
#| class-output: bash
#| echo: false
cat(paste0("touch /proj/​",upc,"/completed/username"))
```

- ​Unless you got some kind of error message you should now be finished. To make
sure the file was created you can type

```{r}
#| attr-output: "filename='bash'"
#| class-output: bash
#| echo: false
cat(paste0("ls /proj/​", upc, "/nobackup/username"))
```

- ​It should write out the name of the file if the file exists. ​If you get an error message, please contact us.
