---
title: "Seurat"
author: Åsa Björklund  &  Paulo Czarnewski
date: August 22, 2019
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    highlight: tango
---

```{r setup, include=FALSE}
#define path to this script (wherever it is)
initial.options <- commandArgs(trailingOnly = FALSE)
script_path <- dirname(sub("--file=","",initial.options[grep("--file=",initial.options)]))
script_path <- "/Users/paulo.barenco/Box/repos/workshop-scRNAseq/labs/r_labs/seurat"
setwd(script_path)

text <- readLines("../Knit_labs.Rmd")
l <- function(x, text, lab="SEURAT"){  p = paste0("#",x,"_",lab,":")
                                       return( sub(p,"", text[grep( p ,text)] ))}
```
***
# `r l("DATA",text,"TITLE")`

`r l("DATA",text,"ALL1")`


```{bash, results='hide'}
mkdir data; cd data
curl -O http://cf.10xgenomics.com/samples/cell-exp/3.0.0/pbmc_1k_v2/pbmc_1k_v2_filtered_feature_bc_matrix.h5
curl -O http://cf.10xgenomics.com/samples/cell-exp/3.0.0/pbmc_1k_v3/pbmc_1k_v3_filtered_feature_bc_matrix.h5
curl -O http://cf.10xgenomics.com/samples/cell-exp/3.0.0/pbmc_1k_protein_v3/pbmc_1k_protein_v3_filtered_feature_bc_matrix.h5
cd ..
```

`r l("DATA",text,"ALL2")`

```{r, message='hide',warning='hide'}
suppressMessages(require(Seurat))
suppressMessages(require(scater))
suppressMessages(require(Matrix))
```

`r l("DATA",text,"ALL3")`


```{r,message='hide',warning='hide'}
v3.1k <- Read10X_h5("data/pbmc_1k_v3_filtered_feature_bc_matrix.h5", use.names = T)
v2.1k <- Read10X_h5("data/pbmc_1k_v2_filtered_feature_bc_matrix.h5", use.names = T)
p3.1k <- Read10X_h5("data/pbmc_1k_protein_v3_filtered_feature_bc_matrix.h5", use.names = T)
p3.1k <- p3.1k$`Gene Expression`
```

***
# `r l("OBJ",text,"TITLE")`

`r l("OBJ",text,"ALL1")`

```{r}
sdata.v2.1k <- CreateSeuratObject(v2.1k, project = "v2.1k")
sdata.v3.1k <- CreateSeuratObject(v3.1k, project = "v3.1k")
sdata.p3.1k <- CreateSeuratObject(p3.1k, project = "p3.1k")

# Merge datasets into one single seurat object
alldata <- merge(sdata.v2.1k, c(sdata.v3.1k,sdata.p3.1k), add.cell.ids=c("v2.1k","v3.1k","p3.1k"))

# Add in a metadata column that indicates v2 vs v3 chemistry
alldata$Chemistry <- ifelse(alldata$orig.ident == "v2.1k","v2","v3")
```

`r l("OBJ",text,"ALL2")`

```{r ,  results='hold'}
as.data.frame(alldata@assays$RNA@counts[1:10,1:2])
head(alldata@meta.data,10)
```


***
# `r l("QC",text,"TITLE")[1]`

`r l("QC",text,"ALL1")`

```{r, results='hold'}
# Way1: Doing it using Seurat function
alldata <- PercentageFeatureSet(alldata, "^MT-", col.name = "percent_mito")

# Way2: Doing it manually
total_counts_per_cell <- colSums(  alldata@assays$RNA@counts  )

mito_genes <- rownames(alldata)[grep("^MT-",rownames(alldata))]
head(mito_genes,10)
alldata$percent_mito <- colSums(  alldata@assays$RNA@counts[mito_genes,]  ) / total_counts_per_cell

```

`r l("QC",text,"ALL2")`

```{r, results='hold'}
# Way1: Doing it using Seurat function
alldata <- PercentageFeatureSet(alldata, "^RP[SL]", col.name = "percent_ribo")

# Way2: Doing it manually
ribo_genes <- rownames(alldata)[grep("^RP[SL]",rownames(alldata))]
head(ribo_genes,10)
alldata$percent_ribo <- colSums(  alldata@assays$RNA@counts[ribo_genes,]  ) / total_counts_per_cell
```

***
# `r l("QC",text,"TITLE2")`

`r l("QC",text,"ALL3")`

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=5,fig.width=16}
feats <- c("nFeature_RNA","nCount_RNA","percent_mito","percent_ribo")
cowplot::plot_grid(ncol = 1,
VlnPlot(alldata, group.by= "orig.ident", features = feats, pt.size = 0.1,ncol = 4) + NoLegend(),
VlnPlot(alldata, group.by= "Chemistry", features = feats, pt.size = 0.1,ncol = 4) + NoLegend()
)
```


`r l("QC",text,"ALL4")`

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=14,fig.width=16}
cowplot::plot_grid(
  FeatureScatter(alldata, "nCount_RNA"  , "nFeature_RNA", group.by = "orig.ident"),
  FeatureScatter(alldata, "percent_mito", "nFeature_RNA", group.by = "orig.ident"),
  FeatureScatter(alldata, "percent_ribo", "nFeature_RNA", group.by = "orig.ident"),
  FeatureScatter(alldata, "percent_ribo", "percent_mito", group.by = "orig.ident")
)
```

***
# `r l("FILTERING",text,"TITLE")`

## `r l("FILTERING",text,"TITLE2")`

`r l("FILTERING",text,"ALL1")`

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=5,fig.width=4}
selected <- WhichCells(alldata, expression = percent_mito < 25)
length(selected)

# and subset the object to only keep those cells
data.filt <- subset(alldata, cells = selected)

# plot violins for new data
VlnPlot(data.filt, features = "percent_mito")
```

`r l("FILTERING",text,"ALL2")`

## `r l("FILTERING",text,"TITLE3")`

`r l("FILTERING",text,"ALL3")`


```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=5,fig.width=16}
#start with cells with many genes detected.
high.det.v3 <- WhichCells(data.filt, expression = nFeature_RNA > 4100)
high.det.v2 <- WhichCells(data.filt, expression = nFeature_RNA > 2000 & orig.ident == "v2.1k")

# remove these cells
data.filt <- subset(data.filt, cells=setdiff(WhichCells(data.filt),c(high.det.v2,high.det.v3)))

# check number of cells
ncol(data.filt)
```

`r l("FILTERING",text,"ALL4")`

## `r l("FILTERING",text,"TITLE4")`

`r l("FILTERING",text,"ALL5")`


```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=8,fig.width=16}
feats <- c("nFeature_RNA","nCount_RNA","percent_mito","percent_ribo")
cowplot::plot_grid(ncol = 1,
VlnPlot(data.filt, group.by= "orig.ident", features = feats, pt.size = 0.1,ncol = 4) + NoLegend(),
VlnPlot(data.filt, group.by= "Chemistry", features = feats, pt.size = 0.1,ncol = 4) + NoLegend()
)
```



# `r l("CELLCYCLE",text,"TITLE")`

`r l("CELLCYCLE",text,"ALL1")`

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=4,fig.width=8}
data.filt <- NormalizeData(data.filt)

data.filt <- CellCycleScoring(object = data.filt,
                              g2m.features = cc.genes$g2m.genes,
                              s.features = cc.genes$s.genes)

VlnPlot(data.filt, features = c("S.Score","G2M.Score"), group.by= "orig.ident")
```


`r l("CELLCYCLE",text,"ALL2")`








