---
description: Reduce high-dimensional gene expression data from
subtitle:  Bioconductor Toolkit
title:  Dimensionality Reduction
---

<div>

> **Note**
>
> Code chunks run R commands unless otherwise specified.

</div>

## Data preparation

First, let's load all necessary libraries and the QC-filtered dataset
from the previous step.

``` {r}
suppressPackageStartupMessages({
  library(scater)
  library(scran)
  library(cowplot)
  library(ggplot2)
  library(rafalib)
  library(umap)
})

sce <- readRDS("data/results/covid_qc.rds")
```

## Feature selection

Next, we first need to define which features/genes are important in our
dataset to distinguish cell types. For this purpose, we need to find
genes that are highly variable across cells, which in turn will also
provide a good separation of the cell clusters.

``` {r}
#| fig-height: 4
#| fig-width: 8

sce <- computeSumFactors(sce, sizes=c(20, 40, 60, 80))
sce <- logNormCounts(sce)
var.out <- modelGeneVar(sce, method="loess")
hvgs = getTopHVGs(var.out, n=2000)

mypar(1,2)
#plot mean over TOTAL variance
# Visualizing the fit:
fit.var <- metadata(var.out)
plot(fit.var$mean, fit.var$var, xlab="Mean of log-expression",
    ylab="Variance of log-expression")
curve(fit.var$trend(x), col="dodgerblue", add=TRUE, lwd=2)

#Select 1000 top variable genes
hvg.out <- getTopHVGs(var.out, n=1000)

# highligt those cells in the plot
cutoff <- rownames(var.out) %in% hvg.out
points(fit.var$mean[cutoff], fit.var$var[cutoff], col="red", pch=16,cex=.6)

#plot mean over BIOLOGICAL variance
plot(var.out$mean, var.out$bio, pch=16, cex=0.4, xlab="Mean log-expression", ylab="Variance of log-expression")
lines(c(min(var.out$mean),max(var.out$mean)), c(0,0), col="dodgerblue", lwd=2)
points(var.out$mean[cutoff], var.out$bio[cutoff], col="red", pch=16,cex=.6)
```

## Z-score transformation

Now that the data is prepared, we now proceed with PCA. Since each gene
has a different expression level, it means that genes with higher
expression values will naturally have higher variation that will be
captured by PCA. This means that we need to somehow give each gene a
similar weight when performing PCA (see below). The common practice is
to center and scale each gene before performing PCA. This exact scaling
is called Z-score normalization it is very useful for PCA, clustering
and plotting heatmaps. Additionally, we can use regression to remove any
unwanted sources of variation from the dataset, such as `cell cycle`,
`sequencing depth`, `percent mitochondria`. This is achieved by doing a
generalized linear regression using these parameters as co-variates in
the model. Then the residuals of the model are taken as the *regressed
data*. Although perhaps not in the best way, batch effect regression can
also be done here. By default variables are scaled in the PCA step and
is not done separately. But it could be achieved by running the commands
below:

However, unlike the Seurat, this step is implemented inside the PCA
function below. Here we will show you how to add the scaledData back to
the object.

``` {r}
# sce@assays$data@listData$scaled.data <- apply(exprs(sce)[rownames(hvg.out),,drop=FALSE],2,function(x) scale(x,T,T))
# rownames(sce@assays$data@listData$scaled.data) <- rownames(hvg.out)
```

## PCA

Performing PCA has many useful applications and interpretations, which
much depends on the data used. In the case of life sciences, we want to
segregate samples based on gene expression patterns in the data.

We use the `logcounts` and then set `scale_features` to TRUE in order to
scale each gene.

``` {r}
#runPCA and specify the variable genes to use for dim reduction with subset_row
sce <- runPCA(sce, exprs_values = "logcounts", ncomponents = 50, subset_row = hvg.out, scale = TRUE)
```

We then plot the first principal components.

``` {r}
#| fig-height: 4
#| fig-width: 13

plot_grid(ncol = 3,
  plotReducedDim(sce,dimred = "PCA",colour_by = "sample",ncomponents = 1:2,point_size = 0.6),
  plotReducedDim(sce,dimred = "PCA",colour_by = "sample",ncomponents = 3:4,point_size = 0.6),
  plotReducedDim(sce,dimred = "PCA",colour_by = "sample",ncomponents = 5:6,point_size = 0.6)
)
```

To identify which genes (Seurat) or metadata parameters (Scater/Scran)
contribute the most to each PC, one can retrieve the loading matrix
information. Unfortunately, this is not implemented in Scater/Scran, so
you will need to compute PCA using `logcounts`.

``` {r}
#| fig-height: 4
#| fig-width: 16

plot_grid(ncol = 2, plotExplanatoryPCs(sce))
```

We can also plot the amount of variance explained by each PC.

``` {r}
#| fig-height: 4
#| fig-width: 4

mypar()
plot(attr(reducedDim(sce,"PCA"),"percentVar")[1:50]*100,type="l",ylab="% variance",xlab="Principal component #")
points(attr(reducedDim(sce,"PCA"),"percentVar")[1:50]*100,pch=21,bg="grey",cex=.5)
```

Based on this plot, we can see that the top 8 PCs retain a lot of
information, while other PCs contain progressively less. However, it is
still advisable to use more PCs since they might contain information
about rare cell types (such as platelets and DCs in this dataset)

## tSNE

We can now run [BH-tSNE](https://arxiv.org/abs/1301.3342).

``` {r}
set.seed(42)
sce <- runTSNE(sce, dimred = "PCA", n_dimred = 30, perplexity = 30,name = "tSNE_on_PCA")
```

We can now plot the tSNE colored per dataset. We can clearly see the
effect of batches present in the dataset.

``` {r}
#| fig-height: 5
#| fig-width: 7

plot_grid(ncol = 3,plotReducedDim(sce,dimred = "tSNE_on_PCA",colour_by = "sample"))
```

## UMAP

We can now run [UMAP](https://arxiv.org/abs/1802.03426) for cell
embeddings.

``` {r}
sce <- runUMAP(sce,dimred = "PCA", n_dimred = 30,   ncomponents = 2,name = "UMAP_on_PCA")
#see ?umap and ?runUMAP for more info
```

UMAP is plotted colored per dataset. Although less distinct as in the
tSNE, we still see quite an effect of the different batches in the data.

``` {r}
sce <- runUMAP(sce,dimred = "PCA", n_dimred = 30,   ncomponents = 10, name = "UMAP10_on_PCA")
#see ?umap and ?runUMAP for more info
```

We can now plot PCA, UMAP and tSNE side by side for comparison. Have a
look at the UMAP and tSNE, what similarities/differences do you see, can
you explain the differences based on what you learned during the
lecture? Also, we can conclude from the dimensionality reductions that
our dataset contains a batch effect that needs to be corrected before
proceeding to clustering and differential gene expression analysis.

``` {r}
#| fig-height: 4.5
#| fig-width: 15

plot_grid(ncol = 3,
  plotReducedDim(sce,dimred = "UMAP_on_PCA",colour_by = "sample")+
    ggplot2::ggtitle(label ="UMAP_on_PCA"),
  plotReducedDim(sce,dimred = "UMAP10_on_PCA",colour_by = "sample",ncomponents = 1:2)+
    ggplot2::ggtitle(label ="UMAP10_on_PCA"),
  plotReducedDim(sce,dimred = "UMAP10_on_PCA",colour_by = "sample",ncomponents = 3:4)+
    ggplot2::ggtitle(label ="UMAP10_on_PCA")
)
```

<div>

> **Discuss**
>
> We have now done Variable gene selection, PCA and UMAP with the
> settings we selected for you. Test a few different ways of selecting
> variable genes, number of PCs for UMAP and check how it influences
> your embedding.

</div>

## Z-scores & DR graphs

Although running a second dimensionality reduction (i.e tSNE or UMAP) on
PCA would be a standard approach (because it allows higher computation
efficiency), the options are actually limitless. Below we will show a
couple of other common options such as running directly on the scaled
data (z-scores) (which was used for PCA) or on a graph built from scaled
data. We will show from now on only UMAP, but the same applies for tSNE.

### UMAP from z-scores

To run tSNE or UMAP on the scaled data, one first needs to select the
number of variables to use. This is because including dimensions that do
contribute to the separation of your cell types will in the end mask
those differences. Another reason for it is because running with all
genes/features also will take longer or might be computationally
unfeasible. Therefore we will use the scaled data of the highly variable
genes.

``` {r}
sce <- runUMAP(sce, exprs_values='logcounts',name = "UMAP_on_ScaleData")
```

### UMAP from graph

To run tSNE or UMAP on the a graph, we first need to build a graph from
the data. In fact, both tSNE and UMAP first build a graph from the data
using a specified distance matrix and then optimize the embedding. Since
a graph is just a matrix containing distances from cell to cell and as
such, you can run either UMAP or tSNE using any other distance metric
desired. Euclidean and Correlation are usually the most commonly used.

``` {r}
#Build Graph
nn <- RANN::nn2(reducedDim(sce,"PCA"),k = 30)
names(nn) <- c("idx","dist")
g <- buildKNNGraph(sce, k=30, use.dimred="PCA")
reducedDim(sce,"KNN") <- igraph::as_adjacency_matrix(g)

#Run UMAP and rename it for comparisson
# temp <- umap::umap.defaults
try(reducedDim(sce,"UMAP_on_Graph") <- NULL)
reducedDim(sce,"UMAP_on_Graph") <- uwot::umap(X=NULL, n_components = 2, nn_method=nn)
```

We can now plot the UMAP comparing both on PCA vs ScaledSata vs Graph.

``` {r}
#| fig-height: 4.5
#| fig-width: 15

plot_grid(ncol = 3,
  plotReducedDim(sce, dimred = "UMAP_on_PCA", colour_by = "sample")+
    ggplot2::ggtitle(label ="UMAP_on_PCA"),
  plotReducedDim(sce, dimred = "UMAP_on_ScaleData", colour_by = "sample")+
    ggplot2::ggtitle(label ="UMAP_on_ScaleData"),
  plotReducedDim(sce, dimred = "UMAP_on_Graph", colour_by = "sample")+
    ggplot2::ggtitle(label ="UMAP_on_Graph")
)
```

## Genes of interest

Let's plot some marker genes for different cell types onto the
embedding.

  Markers                    Cell Type
  -------------------------- -------------------
  CD3E                       T cells
  CD3E CD4                   CD4+ T cells
  CD3E CD8A                  CD8+ T cells
  GNLY, NKG7                 NK cells
  MS4A1                      B cells
  CD14, LYZ, CST3, MS4A7     CD14+ Monocytes
  FCGR3A, LYZ, CST3, MS4A7   FCGR3A+ Monocytes
  FCER1A, CST3               DCs

``` {r}
#| fig-height: 16
#| fig-width: 12

plotlist <- list()
for(i in c("CD3E","CD4","CD8A","NKG7","GNLY","MS4A1","CD14","LYZ","MS4A7","FCGR3A","CST3","FCER1A")){
  plotlist[[i]] <- plotReducedDim(sce,dimred = "UMAP_on_PCA",colour_by = i,by_exprs_values = "logcounts") +
  scale_fill_gradientn(colours = colorRampPalette(c("grey90","orange3","firebrick","firebrick","red","red" ))(10)) +
  ggtitle(label = i)+ theme(plot.title = element_text(size=20)) }
plot_grid(ncol=3, plotlist = plotlist)
```

<div>

> **Discuss**
>
> Select some of your dimensionality reductions and plot some of the QC
> stats that were calculated in the previous lab. Can you see if some of
> the separation in your data is driven by quality of the cells?

</div>

## Save data

We can finally save the object for use in future steps.

``` {r}
path <- "data/results/covid_qc_dm.rds"
if(!file.exists(path)) saveRDS(sce,path)
```

## Session info

```{=html}
<details>
```
```{=html}
<summary>
```
Click here
```{=html}
</summary>
```
``` {r}
sessionInfo()
```

```{=html}
</details>
```
