---
title: "Scater/Scran: Dimensionality reduction"
#CSS_ALL:
---

#DIMRED_MAIN:

#CHUNK_OPT:

#DIMRED_TITLE:

#DIMRED_ALL1:

```{r, message='hide',warning='hide',results='hold'}
suppressPackageStartupMessages({
  library(scater)
  library(scran)
  library(cowplot)
  library(ggplot2)
  library(rafalib)
  library(umap)
})

sce <- readRDS("data/3pbmc_qc.rds")
```

#DIMRED_ALL1.1:

#DIMRED_ALL2:

```{r}
sce <- computeSumFactors(sce, sizes=c(20, 40, 60, 80))
sce <- normalize(sce)
var.fit <- trendVar(sce, use.spikes=FALSE,method="loess",loess.args=list(span=0.02))
var.out <- decomposeVar(sce, var.fit)

mypar(1,2)
#plot mean over TOTAL variance
plot(var.out$mean, var.out$total, pch=16, cex=0.4, xlab="Mean log-expression",
     ylab="Variance of log-expression")
o <- order(var.out$mean)
lines(var.out$mean[o], var.out$tech[o], col="dodgerblue", lwd=2)

cutoff_value <- 0.2
cutoff <- var.out$bio > cutoff_value
points(var.out$mean[cutoff], var.out$total[cutoff], col="red", pch=16,cex=.6)

#plot mean over BIOLOGICAL variance
plot(var.out$mean, var.out$bio, pch=16, cex=0.4, xlab="Mean log-expression",
     ylab="Variance of log-expression")
lines(c(min(var.out$mean),max(var.out$mean)), c(0,0), col="dodgerblue", lwd=2)
points(var.out$mean[cutoff], var.out$bio[cutoff], col="red", pch=16,cex=.6)

hvg.out <- var.out[which(var.out$FDR <= 0.05 & var.out$bio >= cutoff_value),]
hvg.out <- hvg.out[order(hvg.out$bio, decreasing=TRUE),]

print(nrow(hvg.out))
```

#DIMRED_ALL3:

#DIMRED_ALL4:

#DIMRED_SCATER4:

```{r}
# sce@assays$data@listData$scaled.data <- apply(exprs(sce)[rownames(hvg.out),,drop=FALSE],2,function(x) scale(x,T,T))
# rownames(sce@assays$data@listData$scaled.data) <- rownames(hvg.out)
```


#PCA_TITLE:

#PCA_ALL1:

#PCA_SCRAN:

```{r}
#Default Scater way
sce <- runPCA(sce, exprs_values = "logcounts",  scale_features = T,
              ncomponents = 30, feature_set = rownames(hvg.out),method = "prcomp")

#For some reason Scater removes the dimnames of "logcounts" after PCA, so we put it back
dimnames(sce@assays$data@listData$logcounts) <- dimnames(sce@assays$data@listData$counts)

#2nd way:
#sce <- runPCA(sce, exprs_values = "scaled.data", scale_features = FALSE,
#              ncomponents = 30, feature_set = rownames(hvg.out) )
```

#PCA_ALL2:

```{r, fig.asp=.28}
plot_grid(ncol = 3,
  plotReducedDim(sce,use_dimred = "PCA",colour_by = "sample_id",ncomponents = 1:2,add_ticks = F, point_size = 0.6),
  plotReducedDim(sce,use_dimred = "PCA",colour_by = "sample_id",ncomponents = 3:4,add_ticks = F, point_size = 0.6),
  plotReducedDim(sce,use_dimred = "PCA",colour_by = "sample_id",ncomponents = 5:6,add_ticks = F, point_size = 0.6) )

```

#PCA_ALL3:

```{r,fig.asp=.4}
plot_grid(ncol = 2, plotExplanatoryPCs(sce))
```

#PCA_ALL5:

```{r,fig.width=5,fig.height=5}
mypar()
plot(attr(sce@reducedDims$PCA,"percentVar")[1:50]*100,type="l",ylab="% variance",xlab="Principal component #")
points(attr(sce@reducedDims$PCA,"percentVar")[1:50]*100,pch=21,bg="grey",cex=.5)
```

#PCA_ALL6:

#tSNE_TITLE:

#tSNE_ALL1:

```{r, fig.asp=1}
set.seed(42)
sce <- runTSNE(sce, use_dimred = "PCA", n_dimred = 30, 
               perplexity = 30)
#see ?Rtsne and ?runTSNE for more info
reducedDimNames(sce)[reducedDimNames(sce)=="TSNE"] <- "tSNE_on_PCA"
```

#tSNE_ALL2:

```{r,fig.asp=.28}
plot_grid(ncol = 3,plotReducedDim(sce,use_dimred = "tSNE_on_PCA",colour_by = "sample_id",add_ticks = F))
```


#UMAP_TITLE:

#UMAP_ALL1:

```{r}
sce <- runUMAP(sce,use_dimred = "PCA", n_dimred = 30,   ncomponents = 2)

#We need to rename it to not overide with other UMAP computations
try(sce@reducedDims$UMAP_on_RNA <- NULL)
reducedDimNames(sce)[reducedDimNames(sce)=="UMAP"] <- "UMAP_on_PCA"
#see ?umap and ?runUMAP for more info
```

#UMAP_ALL2:

```{r}
sce <- runUMAP(sce,use_dimred = "PCA", n_dimred = 30,   ncomponents = 10)
#see ?umap and ?runUMAP for more info

#We need to rename it to not overide with other UMAP computations
try(sce@reducedDims$UMAP10_on_RNA <- NULL)
reducedDimNames(sce)[reducedDimNames(sce)=="UMAP"] <- "UMAP10_on_PCA"
```

#UMAP_ALL2.1:

```{r,fig.asp=.28}
plot_grid(ncol = 3,
          plotReducedDim(sce,use_dimred = "UMAP_on_PCA",colour_by = "sample_id",add_ticks = F)+
            ggplot2::ggtitle(label ="UMAP_on_PCA"),
          plotReducedDim(sce,use_dimred = "UMAP10_on_PCA",colour_by = "sample_id",ncomponents = 1:2,add_ticks = F)+
            ggplot2::ggtitle(label ="UMAP10_on_PCA"),
          plotReducedDim(sce,use_dimred = "UMAP10_on_PCA",colour_by = "sample_id",ncomponents = 3:4,add_ticks = F)+
            ggplot2::ggtitle(label ="UMAP10_on_PCA")
)
```


#DIMRED_TITLE2:

#DIMRED_ALL5:

#DIMRED_ALL5.0:

#DIMRED_ALL5.1:

```{r,fig.asp=.28}
sce <- runUMAP(sce, exprs_values='logcounts', feature_set = rownames(hvg.out))

#We need to rename it to not overide with other UMAP computations
try(sce@reducedDims$UMAP_on_ScaleData <- NULL)
reducedDimNames(sce)[reducedDimNames(sce)=="UMAP"] <- "UMAP_on_ScaleData"
```

#DIMRED_ALL5.2:

#DIMRED_ALL5.2.1:

```{r,fig.asp=.28}
#Build Graph
g <- buildKNNGraph(sce,k=30,use.dimred="PCA",assay.type="RNA")
sce@reducedDims$KNN <- igraph::as_adjacency_matrix(g)


#Run UMAP and rename it for comparisson
# temp <- umap::umap.defaults
# temp$input <- "dist"
sce <- runUMAP(sce,use_dimred = "KNN", ncomponents = 2, input="data")
try(sce@reducedDims$UMAP_on_Graph <- NULL)
reducedDimNames(sce)[reducedDimNames(sce)=="UMAP"] <- "UMAP_on_Graph"
```


#DIMRED_ALL5.3:

```{r,fig.asp=.28}
plot_grid(ncol = 3,
  plotReducedDim(sce, use_dimred = "UMAP_on_PCA", colour_by = "sample_id",add_ticks = F)+ 
    ggplot2::ggtitle(label ="UMAP_on_PCA"),
  plotReducedDim(sce, use_dimred = "UMAP_on_ScaleData", colour_by = "sample_id",add_ticks = F)+
    ggplot2::ggtitle(label ="UMAP_on_ScaleData"),
  plotReducedDim(sce, use_dimred = "UMAP_on_Graph", colour_by = "sample_id",add_ticks = F)+
    ggplot2::ggtitle(label ="UMAP_on_Graph")
)
```

#DIMRED_TITLE3:


#DIMRED_ALL6:

#MARKER_TABLE:

```{r,fig.asp=.9}
plotlist <- list()
for(i in c("CD3E","CD4","CD8A","NKG7","GNLY","MS4A1","CD14","LYZ","MS4A7","FCGR3A","CST3","FCER1A")){
  plotlist[[i]] <- plotReducedDim(sce,use_dimred = "UMAP_on_PCA",colour_by = i,by_exprs_values = "logcounts",add_ticks = F) +
  scale_fill_gradientn(colours = colorRampPalette(c("grey90","orange3","firebrick","firebrick","red","red" ))(10)) +
  ggtitle(label = i)+ theme(plot.title = element_text(size=20)) }
plot_grid(ncol=3, plotlist = plotlist)
```


#DIMRED_ALL7:

```{r}
saveRDS(sce,"data/3pbmc_qc_dm.rds")
```

#SESSION_INFO:

```{r}
sessionInfo()
```
















