---
title: "Scater/Scran: Dimensionality reduction"
#CSS_ALL:
editor_options:
  chunk_output_type: console
---

#DIMRED_MAIN:

#CHUNK_OPT:

#DIMRED_TITLE:

#DIMRED_ALL1:

```{r, message='hide',warning='hide',results='hold'}
suppressPackageStartupMessages({
  library(scater)
  library(scran)
  library(cowplot)
  library(ggplot2)
  library(rafalib)
  library(umap)
})

sce <- readRDS("data/results/covid_qc.rds")
```

#DIMRED_ALL1.1:

#DIMRED_ALL2:

```{r}
sce <- computeSumFactors(sce, sizes=c(20, 40, 60, 80))
sce <- logNormCounts(sce)
var.out <- modelGeneVar(sce, method="loess")
hvgs = getTopHVGs(var.out, n=2000)

mypar(1,2)
#plot mean over TOTAL variance
# Visualizing the fit:
fit.var <- metadata(var.out)
plot(fit.var$mean, fit.var$var, xlab="Mean of log-expression",
    ylab="Variance of log-expression")
curve(fit.var$trend(x), col="dodgerblue", add=TRUE, lwd=2)

#Select 1000 top variable genes
hvg.out <- getTopHVGs(var.out, n=1000)

# highligt those cells in the plot
cutoff <- rownames(var.out) %in% hvg.out
points(fit.var$mean[cutoff], fit.var$var[cutoff], col="red", pch=16,cex=.6)


#plot mean over BIOLOGICAL variance
plot(var.out$mean, var.out$bio, pch=16, cex=0.4, xlab="Mean log-expression",
     ylab="Variance of log-expression")
lines(c(min(var.out$mean),max(var.out$mean)), c(0,0), col="dodgerblue", lwd=2)
points(var.out$mean[cutoff], var.out$bio[cutoff], col="red", pch=16,cex=.6)

```

#DIMRED_ALL3:

#DIMRED_ALL4:

#DIMRED_SCATER4:

```{r}
# sce@assays$data@listData$scaled.data <- apply(exprs(sce)[rownames(hvg.out),,drop=FALSE],2,function(x) scale(x,T,T))
# rownames(sce@assays$data@listData$scaled.data) <- rownames(hvg.out)
```


#PCA_TITLE:

#PCA_ALL1:

#PCA_SCRAN:

```{r}
#runPCA and specify the variable genes to use for dim reduction with subset_row
sce <- runPCA(sce, exprs_values = "logcounts",
              ncomponents = 50,
              subset_row = hvg.out,
              scale = TRUE)
```

#PCA_ALL2:

```{r, fig.asp=.28}
plot_grid(ncol = 3,
  plotReducedDim(sce,dimred = "PCA",colour_by = "sample",ncomponents = 1:2,point_size = 0.6),
  plotReducedDim(sce,dimred = "PCA",colour_by = "sample",ncomponents = 3:4,point_size = 0.6),
  plotReducedDim(sce,dimred = "PCA",colour_by = "sample",ncomponents = 5:6,point_size = 0.6) )

```

#PCA_ALL3:

```{r,fig.asp=.4}
plot_grid(ncol = 2, plotExplanatoryPCs(sce))
```

#PCA_ALL5:

```{r,fig.width=5,fig.height=5}
mypar()
plot(attr(reducedDim(sce,"PCA"),"percentVar")[1:50]*100,type="l",ylab="% variance",xlab="Principal component #")
points(attr(reducedDim(sce,"PCA"),"percentVar")[1:50]*100,pch=21,bg="grey",cex=.5)
```

#PCA_ALL6:

#tSNE_TITLE:

#tSNE_ALL1:

```{r, fig.asp=1}
set.seed(42)
sce <- runTSNE(sce, dimred = "PCA", n_dimred = 30,
               perplexity = 30,name = "tSNE_on_PCA")
```

#tSNE_ALL2:

```{r,fig.asp=.28}
plot_grid(ncol = 3,plotReducedDim(sce,dimred = "tSNE_on_PCA",colour_by = "sample"))
```


#UMAP_TITLE:

#UMAP_ALL1:

```{r}
sce <- runUMAP(sce,dimred = "PCA", n_dimred = 30,   ncomponents = 2,name = "UMAP_on_PCA")
#see ?umap and ?runUMAP for more info
```

#UMAP_ALL2:

```{r}
sce <- runUMAP(sce,dimred = "PCA", n_dimred = 30,   ncomponents = 10, name = "UMAP10_on_PCA")
#see ?umap and ?runUMAP for more info
```

#UMAP_ALL2.1:

```{r,fig.asp=.28}
plot_grid(ncol = 3,
          plotReducedDim(sce,dimred = "UMAP_on_PCA",colour_by = "sample")+
            ggplot2::ggtitle(label ="UMAP_on_PCA"),
          plotReducedDim(sce,dimred = "UMAP10_on_PCA",colour_by = "sample",ncomponents = 1:2)+
            ggplot2::ggtitle(label ="UMAP10_on_PCA"),
          plotReducedDim(sce,dimred = "UMAP10_on_PCA",colour_by = "sample",ncomponents = 3:4)+
            ggplot2::ggtitle(label ="UMAP10_on_PCA")
)
```


#DIMRED_TITLE2:

#DIMRED_ALL5:

#DIMRED_ALL5.0:

#DIMRED_ALL5.1:

```{r,fig.asp=.28}
sce <- runUMAP(sce, exprs_values='logcounts',name = "UMAP_on_ScaleData")
```

#DIMRED_ALL5.2:

#DIMRED_ALL5.2.1:

```{r,fig.asp=.28}
#Build Graph
nn <- RANN::nn2(reducedDim(sce,"PCA"),k = 30)
names(nn) <- c("idx","dist")
g <- buildKNNGraph(sce, k=30, use.dimred="PCA")
reducedDim(sce,"KNN") <- igraph::as_adjacency_matrix(g)


#Run UMAP and rename it for comparisson
# temp <- umap::umap.defaults
try(reducedDim(sce,"UMAP_on_Graph") <- NULL)
reducedDim(sce,"UMAP_on_Graph") <- uwot::umap(X=NULL, n_components = 2, nn_method=nn)
```


#DIMRED_ALL5.3:

```{r,fig.asp=.28}
plot_grid(ncol = 3,
  plotReducedDim(sce, dimred = "UMAP_on_PCA", colour_by = "sample")+
    ggplot2::ggtitle(label ="UMAP_on_PCA"),
  plotReducedDim(sce, dimred = "UMAP_on_ScaleData", colour_by = "sample")+
    ggplot2::ggtitle(label ="UMAP_on_ScaleData"),
  plotReducedDim(sce, dimred = "UMAP_on_Graph", colour_by = "sample")+
    ggplot2::ggtitle(label ="UMAP_on_Graph")
)
```

#DIMRED_TITLE3:


#DIMRED_ALL6:

#MARKER_TABLE:

```{r,fig.asp=.9}
plotlist <- list()
for(i in c("CD3E","CD4","CD8A","NKG7","GNLY","MS4A1","CD14","LYZ","MS4A7","FCGR3A","CST3","FCER1A")){
  plotlist[[i]] <- plotReducedDim(sce,dimred = "UMAP_on_PCA",colour_by = i,by_exprs_values = "logcounts") +
  scale_fill_gradientn(colours = colorRampPalette(c("grey90","orange3","firebrick","firebrick","red","red" ))(10)) +
  ggtitle(label = i)+ theme(plot.title = element_text(size=20)) }
plot_grid(ncol=3, plotlist = plotlist)
```


#DIMRED_ALL7:

#DIMRED_ALL8:

```{r}
saveRDS(sce,"data/results/covid_qc_dm.rds")
```

#SESSION_INFO:

```{r}
sessionInfo()
```
