---
title: "Seurat: Quality control"
#CSS_ALL:
editor_options: 
  chunk_output_type: console
---

#CHUNK_OPT:

***
#DATA_TITLE:

#DATA_ALL1:

```{r,message='hide',warning='hide',results='hold'}
webpath <- "https://raw.githubusercontent.com/NBISweden/workshop-scRNAseq/new_dataset/labs/data/covid_data_GSE149689/sub/"
dir.create("./data/raw",recursive = T)
file_list <- c("Normal_PBMC_13.h5","Normal_PBMC_14.h5","Normal_PBMC_5.h5",
               "nCoV_PBMC_15.h5","nCoV_PBMC_17.h5","nCoV_PBMC_1.h5")
for(i in file_list){
  download.file( url = paste0(webpath,i) , 
                 destfile = paste0("./data/raw/",i))
}
```


#DATA_ALL2:

```{r, message='hide',warning='hide',results='hold'}
suppressMessages(require(Seurat))
suppressMessages(require(Matrix))
if(!require(DoubletFinder)){
   remotes::install_github('chris-mcginnis-ucsf/DoubletFinder',upgrade = FALSE, dependencies=FALSE)
}
suppressMessages(require(DoubletFinder))
```

#DATA_ALL3:

```{r,message='hide',warning='hide',results='hold'}
cov.15 <- Seurat::Read10X_h5(
  filename = "data/raw/nCoV_PBMC_15.h5",
  use.names = T)
cov.1 <- Seurat::Read10X_h5(
  filename = "data/raw/nCoV_PBMC_1.h5",
  use.names = T)
cov.17 <- Seurat::Read10X_h5(
  filename = "data/raw/nCoV_PBMC_17.h5",
  use.names = T)

ctrl.5 <- Seurat::Read10X_h5(
  filename = "data/raw/Normal_PBMC_5.h5",
  use.names = T)
ctrl.13 <- Seurat::Read10X_h5(
  filename = "data/raw/Normal_PBMC_13.h5",
  use.names = T)
ctrl.14 <- Seurat::Read10X_h5(
  filename = "data/raw/Normal_PBMC_14.h5",
  use.names = T)


```

***
#OBJ_TITLE:

#OBJ_ALL1:

```{r}
sdata.cov15 <- CreateSeuratObject(cov.15,  project = "covid_15")
sdata.cov1 <- CreateSeuratObject(cov.1,  project = "covid_1")
sdata.cov17 <- CreateSeuratObject(cov.17,  project = "covid_17")
sdata.ctrl5 <- CreateSeuratObject(ctrl.5,  project = "ctrl_5")
sdata.ctrl13 <- CreateSeuratObject(ctrl.13,  project = "ctrl_13")
sdata.ctrl14 <- CreateSeuratObject(ctrl.14,  project = "ctrl_14")

# add metadata
sdata.cov1$type = "Covid"
sdata.cov15$type = "Covid"
sdata.cov17$type = "Covid"
sdata.ctrl5$type = "Ctrl"
sdata.ctrl13$type = "Ctrl"
sdata.ctrl14$type = "Ctrl"



# Merge datasets into one single seurat object
alldata <- merge(sdata.cov15, c(sdata.cov1, sdata.cov17, sdata.ctrl5, sdata.ctrl13, sdata.ctrl14), add.cell.ids=c("covid_15","covid_1","covid_17","ctrl_5","ctrl_13", "ctrl_14"))

```

#OBJ_ALL1.1:

```{r}
# remove all objects that will not be used.
rm(cov.15, cov.1, cov.17, ctrl.5, ctrl.13, ctrl.14, sdata.cov15, sdata.cov1, sdata.cov17, sdata.ctrl5, sdata.ctrl13, sdata.ctrl14)

# run garbage collect to free up memory
gc()
```
#OBJ_ALL2:

```{r ,  results='hold'}
as.data.frame(alldata@assays$RNA@counts[1:10,1:2])
head(alldata@meta.data,10)
```


***
#QC_TITLE:

#QC_ALL1:

#QC_ALL1.1:

```{r, results='hold'}
# Way1: Doing it using Seurat function
alldata <- PercentageFeatureSet(alldata, "^MT-", col.name = "percent_mito")

# Way2: Doing it manually
total_counts_per_cell <- colSums(  alldata@assays$RNA@counts  )
mito_genes <- rownames(alldata)[grep("^MT-",rownames(alldata))]
alldata$percent_mito <- colSums(  alldata@assays$RNA@counts[mito_genes,]  ) / total_counts_per_cell

head(mito_genes,10)
```

#QC_ALL2:

```{r, results='hold'}
# Way1: Doing it using Seurat function
alldata <- PercentageFeatureSet(alldata, "^RP[SL]", col.name = "percent_ribo")

# Way2: Doing it manually
ribo_genes <- rownames(alldata)[grep("^RP[SL]",rownames(alldata))]
head(ribo_genes,10)
alldata$percent_ribo <- colSums(  alldata@assays$RNA@counts[ribo_genes,]  ) / total_counts_per_cell
```

#QC_ALL2.2:

```{r, results='hold'}
# Percentage hemoglobin genes - includes all genes starting with HB except HBP.
alldata <- PercentageFeatureSet(alldata, "^HB[^(P)]", col.name = "percent_hb")

alldata <- PercentageFeatureSet(alldata, "PECAM1|PF4", col.name = "percent_plat")

```

***
#QC_TITLE2:

#QC_ALL3:

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=8,fig.width=16}
feats <- c("nFeature_RNA","nCount_RNA","percent_mito","percent_ribo", "percent_hb")
VlnPlot(alldata, group.by= "orig.ident", features = feats, pt.size = 0.1,ncol = 3) + NoLegend()
```

#QC_ALL4:

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=5,fig.width=16}

FeatureScatter(alldata, "nCount_RNA"  , "nFeature_RNA", group.by = "orig.ident", pt.size = .5)

```

#QC_ALL4.4:

***
#FILTERING_TITLE:

#FILTERING_TITLE1:

#FILTERING_ALL0:

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=5,fig.width=4}
selected_c <- WhichCells(alldata, expression = nFeature_RNA > 200)
selected_f <- rownames(alldata)[ Matrix::rowSums(alldata) > 3]

data.filt <- subset(alldata, features=selected_f, cells=selected_c)
dim(data.filt)
```

#FILTERING_ALL3:

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=5,fig.width=16}
# skip for now and run DoubletFinder first!

#high.det.v3 <- WhichCells(data.filt, expression = nFeature_RNA > 4100)
#high.det.v2 <- WhichCells(data.filt, expression = nFeature_RNA > 2000 & orig.ident == "v2.1k")

# remove these cells
#data.filt <- subset(data.filt, cells=setdiff(WhichCells(data.filt),c(high.det.v2,high.det.v3)))

# check number of cells
ncol(data.filt)
```

#FILTERING_ALL01:

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=8,fig.width=12}
#Compute the relative expression of each gene per cell
#Use sparse matrix operations, if your dataset is large, doing matrix devisions the regular way will take a very long time.
par(mar=c(4,8,2,1))
C <- data.filt@assays$RNA@counts
C <-  Matrix::t( Matrix::t(C) / Matrix::colSums(C) ) * 100
most_expressed <- order(apply(C,1,median),decreasing = T)[20:1]
boxplot( as.matrix(t(C[most_expressed,])),cex=.1, las=1, xlab="% total count per cell",col=scales::hue_pal()(20)[20:1],horizontal=TRUE)

```

#FILTERING_ALL02:

#FILTERING_TITLE2:

#FILTERING_ALL1:

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=5,fig.width=16}
selected_mito <- WhichCells(data.filt, expression = percent_mito < 0.20)
selected_ribo <- WhichCells(data.filt, expression = percent_ribo > 0.05)

# and subset the object to only keep those cells
data.filt <- subset(data.filt, cells = selected_mito)
data.filt <- subset(data.filt, cells = selected_ribo)

dim(data.filt)

table(data.filt$orig.ident)
```

#FILTERING_ALL2:

#FILTERING_TITLE4:

#FILTERING_ALL5:

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=8,fig.width=16}
feats <- c("nFeature_RNA","nCount_RNA","percent_mito","percent_ribo", "percent_hb")

VlnPlot(data.filt, group.by= "orig.ident", features = feats, pt.size = 0.1,ncol = 3) + NoLegend()
```

#FILTERING_TITLE5:

#FILTERING_ALL6:

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=4,fig.width=4}
dim(data.filt)

# Filter MALAT1
data.filt <- data.filt[ ! grepl("MALAT1", rownames(data.filt)), ]

# Filter Mitocondrial
data.filt <- data.filt[ ! grepl("^MT-", rownames(data.filt)), ]

# Filter Ribossomal gene (optional if that is a problem on your data)
# data.filt <- data.filt[ ! grepl("^RP[SL]", rownames(data.filt)), ]

# Filter Hemoglobin gene (optional if that is a problem on your data)
data.filt <- data.filt[ ! grepl("^HB[^(P)]", rownames(data.filt)), ]

dim(data.filt)
```



#SEX_TITLE:

#SEX_ALL1:

#SEX_ALL2:

```{r}
genes.file = "data/results/genes.table.csv"

if (!file.exists(genes.file)){
  suppressMessages(require(biomaRt))

  # initialize connection to mart, may take some time if the sites are unresponsive.
  mart <- useMart("ENSEMBL_MART_ENSEMBL", dataset="hsapiens_gene_ensembl")

  # fetch chromosome info plus some other annotations
  genes.table <- try(biomaRt::getBM(attributes = c("ensembl_gene_id",
              "external_gene_name", "description","gene_biotype", "chromosome_name","start_position"),
              mart = mart, useCache = F))
  
  if(!dir.exists("data/results")){dir.create("data/results")}
  if(is.data.frame(genes.table)){write.csv(genes.table, file = genes.file)}
  
  if (!file.exists(genes.file)){
  download.file("https://raw.githubusercontent.com/NBISweden/workshop-scRNAseq/master/labs/misc/genes.table.csv",destfile = "data/results/genes.table.csv")
    genes.table = read.csv(genes.file)
    }

} else {
  genes.table = read.csv(genes.file)
}

genes.table <- genes.table[genes.table$external_gene_name %in% rownames(data.filt),]

```

#SEX_ALL3:

```{r}
chrY.gene = genes.table$external_gene_name[genes.table$chromosome_name == "Y"]

data.filt$pct_chrY = colSums(data.filt@assays$RNA@counts[chrY.gene,]) / colSums(data.filt@assays$RNA@counts)

```

#SEX_ALL4:

```{r}
FeatureScatter(data.filt, feature1 = "XIST", feature2 = "pct_chrY")
```

#SEX_ALL5:

```{r}
VlnPlot(data.filt, features = c("XIST", "pct_chrY"))
```

#SEX_ALL6:


#CELLCYCLE_TITLE:

#CELLCYCLE_ALL1:

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=8,fig.width=8}
# Before running CellCycleScoring the data need to be normalized and logtransformed.
data.filt = NormalizeData(data.filt)


data.filt <- CellCycleScoring(object = data.filt,
                              g2m.features = cc.genes$g2m.genes,
                              s.features = cc.genes$s.genes)
```

#CELLCYCLE_ALL2:

```{r,message='hide',warning='hide', results='hold',results='hold',fig.height=5,fig.width=16}
VlnPlot(data.filt, features = c("S.Score","G2M.Score"), group.by= "orig.ident",ncol = 4, pt.size = .1)
```

#CELLCYCLE_ALL3:


#DOUBLET_TITLE:

#DOUBLET_ALL1:

#DOUBLET_SEURAT1.1:


```{r}
suppressMessages(require(DoubletFinder))

data.filt = FindVariableFeatures(data.filt, verbose = F)
data.filt = ScaleData(data.filt, vars.to.regress = c("nFeature_RNA", "percent_mito"), verbose = F)
data.filt = RunPCA(data.filt, verbose = F, npcs = 20)
data.filt = RunUMAP(data.filt, dims = 1:10, verbose = F)

```

#DOUBLET_SEURAT1.2:

```{r doubletfinder, message=FALSE}
# Can run parameter optimization with paramSweep

#sweep.res <- paramSweep_v3(data.filt)
#sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
#bcmvn <- find.pK(sweep.stats)
#barplot(bcmvn$BCmetric, names.arg = bcmvn$pK, las=2)

# define the expected number of doublet cellscells.
nExp <- round(ncol(data.filt)* 0.04) # expect 4% doublets
data.filt <- doubletFinder_v3(data.filt, pN=0.25, pK = 0.09, nExp = nExp, PCs = 1:10)
```

```{r}

# name of the DF prediction can change, so extract the correct column name.
DF.name = colnames(data.filt@meta.data)[grepl("DF.classification", colnames(data.filt@meta.data))]


cowplot::plot_grid( ncol = 2,
DimPlot(data.filt, group.by = "orig.ident") + NoAxes(),
DimPlot(data.filt, group.by = DF.name) + NoAxes()
)
```


#DOUBLET_ALL1.1:

```{r}
VlnPlot(data.filt, features = "nFeature_RNA", group.by = DF.name, pt.size = .1)

```

#DOUBLET_ALL2:

```{r}
data.filt = data.filt[,data.filt@meta.data[,DF.name] == "Singlet"]
dim(data.filt)
```

#DOUBLET_ALL3:

```{r save,message='hide',warning='hide', results='hold',results='hold',fig.height=5,fig.width=16}

dir.create('data/results', showWarnings = F)

saveRDS(data.filt,"data/results/seurat_covid_qc.rds")
```


#SESSION_INFO:

```{r}
sessionInfo()
```




