---
title: "{{< meta clust_title >}}"
subtitle: "{{< meta subtitle_seurat >}}"
description: "{{< meta clust_description >}}"
format: html
---

::: {.callout-note}
Code chunks run R commands unless otherwise specified.
:::

{{< meta clust_1 >}}

{{< meta clust_2 >}}

```{r}
suppressPackageStartupMessages({
    library(Seurat)
    library(patchwork)
    library(ggplot2)
    library(pheatmap)
    library(clustree)
})
```

```{r}
# download pre-computed data if missing or long compute
fetch_data <- TRUE

# url for source and intermediate data
path_data <- "https://export.uppmax.uu.se/naiss2023-23-3/workshops/workshop-scrnaseq"
path_file <- "data/covid/results/seurat_covid_qc_dr_int.rds"
if (!dir.exists(dirname(path_file))) dir.create(dirname(path_file), recursive = TRUE)
if (fetch_data && !file.exists(path_file)) download.file(url = file.path(path_data, "covid/results/seurat_covid_qc_dr_int.rds"), destfile = path_file)
alldata <- readRDS(path_file)
print(names(alldata@reductions))
```

## {{< meta clust_graphclust >}}

{{< meta clust_graphclust_1 >}}

### {{< meta clust_graphclust_knn >}}

{{< meta clust_graphclust_knn_1 >}}

As we can see above, the **Seurat** function `FindNeighbors()` already computes both the KNN and SNN graphs, in which we can control the minimal percentage of shared neighbours to be kept. See `?FindNeighbors` for additional options.

```{r}
# check if CCA is still the active assay
alldata@active.assay

# set the correct assay.
alldata@active.assay <- "CCA"

alldata <- FindNeighbors(alldata, dims = 1:30, k.param = 60, prune.SNN = 1 / 15)

# check the names for graphs in the object.
names(alldata@graphs)
```

{{< meta clust_graphclust_knn_2 >}}

```{r}
#| fig-height: 6
#| fig-width: 6

pheatmap(alldata@graphs$CCA_nn[1:200, 1:200],
    col = c("white", "black"), border_color = "grey90", main = "KNN graph",
    legend = F, cluster_rows = F, cluster_cols = F, fontsize = 2
)

pheatmap(alldata@graphs$CCA_snn[1:200, 1:200],
    col = colorRampPalette(c("white", "yellow", "red"))(100),
    border_color = "grey90", main = "SNN graph",
    legend = F, cluster_rows = F, cluster_cols = F, fontsize = 2
)
```

### {{< meta clust_graphclust_graph >}}

{{< meta clust_graphclust_graph_1 >}}

In **Seurat**, the function `FindClusters()` will do a graph-based clustering using "Louvain" algorithim by default (`algorithm = 1`). To use the leiden algorithm, you need to set it to `algorithm = 4`. See `?FindClusters` for additional options.

By default it will run clustering on the SNN graph we created in the previous step, but you can also specify different graphs for clustering with `graph.name`. 

```{r}
#| fig-height: 4
#| fig-width: 12
#| results: hide

# Clustering with louvain (algorithm 1) and a few different resolutions
for (res in c(0.1, 0.25, .5, 1, 1.5, 2)) {
    alldata <- FindClusters(alldata, graph.name = "CCA_snn", resolution = res, algorithm = 1)
}

# each time you run clustering, the data is stored in meta data columns:
# seurat_clusters - lastest results only
# CCA_snn_res.XX - for each different resolution you test.
```

```{r}
#| fig-height: 4
#| fig-width: 13

wrap_plots(
    DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.0.5") + ggtitle("louvain_0.5"),
    DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.1") + ggtitle("louvain_1"),
    DimPlot(alldata, reduction = "umap", group.by = "CCA_snn_res.2") + ggtitle("louvain_2"),
    ncol = 3
)
```

{{< meta clust_graphclust_graph_2 >}}

```{r}
#| fig-height: 8
#| fig-width: 8

suppressPackageStartupMessages(library(clustree))
clustree(alldata@meta.data, prefix = "CCA_snn_res.")
```

## {{< meta clust_kmean >}}

{{< meta clust_kmean_1 >}}

```{r}
#| fig-height: 4
#| fig-width: 13

for (k in c(5, 7, 10, 12, 15, 17, 20)) {
    alldata@meta.data[, paste0("kmeans_", k)] <- kmeans(x = alldata@reductions[["pca"]]@cell.embeddings, centers = k, nstart = 100)$cluster
}

wrap_plots(
    DimPlot(alldata, reduction = "umap", group.by = "kmeans_5") + ggtitle("kmeans_5"),
    DimPlot(alldata, reduction = "umap", group.by = "kmeans_10") + ggtitle("kmeans_10"),
    DimPlot(alldata, reduction = "umap", group.by = "kmeans_15") + ggtitle("kmeans_15"),
    ncol = 3
)
```

```{r}
#| fig-height: 8
#| fig-width: 8

clustree(alldata@meta.data, prefix = "kmeans_")
```

## {{< meta clust_hier >}}

### {{< meta clust_hier_dist >}}

{{< meta clust_hier_dist_1 >}}

```{r}
d <- dist(alldata@reductions[["pca"]]@cell.embeddings, method = "euclidean")
```

{{< meta clust_hier_dist_2 >}}

```{r}
# Compute sample correlations
sample_cor <- cor(Matrix::t(alldata@reductions[["pca"]]@cell.embeddings))

# Transform the scale from correlations
sample_cor <- (1 - sample_cor) / 2

# Convert it to a distance object
d2 <- as.dist(sample_cor)
```

### {{< meta clust_hier_clust >}}

{{< meta clust_hier_clust_1 >}}

```{r}
# euclidean
h_euclidean <- hclust(d, method = "ward.D2")

# correlation
h_correlation <- hclust(d2, method = "ward.D2")
```

{{< meta clust_hier_clust_2 >}}

```{r}
#| fig-height: 8
#| fig-width: 14

# euclidean distance
alldata$hc_euclidean_5 <- cutree(h_euclidean, k = 5)
alldata$hc_euclidean_10 <- cutree(h_euclidean, k = 10)
alldata$hc_euclidean_15 <- cutree(h_euclidean, k = 15)

# correlation distance
alldata$hc_corelation_5 <- cutree(h_correlation, k = 5)
alldata$hc_corelation_10 <- cutree(h_correlation, k = 10)
alldata$hc_corelation_15 <- cutree(h_correlation, k = 15)

wrap_plots(
    DimPlot(alldata, reduction = "umap", group.by = "hc_euclidean_5") + ggtitle("hc_euc_5"),
    DimPlot(alldata, reduction = "umap", group.by = "hc_euclidean_10") + ggtitle("hc_euc_10"),
    DimPlot(alldata, reduction = "umap", group.by = "hc_euclidean_15") + ggtitle("hc_euc_15"),
    DimPlot(alldata, reduction = "umap", group.by = "hc_corelation_5") + ggtitle("hc_cor_5"),
    DimPlot(alldata, reduction = "umap", group.by = "hc_corelation_10") + ggtitle("hc_cor_10"),
    DimPlot(alldata, reduction = "umap", group.by = "hc_corelation_15") + ggtitle("hc_cor_15"),
    ncol = 3
) + plot_layout()
```

{{< meta clust_save >}}

```{r}
saveRDS(alldata, "data/covid/results/seurat_covid_qc_dr_int_cl.rds")
```

## {{< meta clust_distribution >}}

{{< meta clust_distribution_1 >}}

Select the **CCA_snn_res.0.5** and plot proportion of samples per cluster and also proportion covid vs ctrl. 

```{r}
#| fig-height: 4
#| fig-width: 9

p1 <- ggplot(alldata@meta.data, aes(x = CCA_snn_res.0.5, fill = orig.ident)) +
    geom_bar(position = "fill")
p2 <- ggplot(alldata@meta.data, aes(x = CCA_snn_res.0.5, fill = type)) +
    geom_bar(position = "fill")

p1 + p2
```

{{< meta clust_distribution_2 >}}

{{< meta clust_distribution_3 >}}

```{r}
ggplot(alldata@meta.data, aes(x = orig.ident, fill = CCA_snn_res.0.5)) +
    geom_bar(position = "fill")
```

:::{.callout-note title="Discuss"}
{{< meta clust_3 >}}
:::

## {{< meta session >}}

<details>
  <summary>Click here</summary>

```{r}
sessionInfo()
```

</details>
