---
title: "{{< meta int_title >}}"
subtitle: "{{< meta subtitle_seurat >}}"
description: "{{< meta int_description >}}"
format: html
---

::: {.callout-note}
Code chunks run R commands unless otherwise specified.
:::

{{< meta int_1 >}}

|Markdown | Language | Library | Ref|
|:---|:---|:---|:---|
|CCA | R | Seurat | [Cell](https://www.sciencedirect.com/science/article/pii/S0092867419305598?via%3Dihub)|
|MNN | R/Python | Scater/Scanpy | [Nat. Biotech.](https://www.nature.com/articles/nbt.4091)|
|Conos | R | conos | [Nat. Methods](https://www.nature.com/articles/s41592-019-0466-z?error=cookies_not_supported&code=5680289b-6edb-40ad-9934-415dac4fdb2f)|
|Conos | R | Harmony | [Nat. Methods](https://www.nature.com/articles/s41592-019-0619-0)|
|Scanorama | Python | scanorama | [Nat. Biotech.](https://www.nature.com/articles/s41587-019-0113-3)|


## {{< meta int_prep >}}

{{< meta int_prep_1 >}}

```{r}
suppressPackageStartupMessages({
    library(Seurat)
    library(ggplot2)
    library(patchwork)
    library(reticulate)
})

# Activate scanorama Python venv
reticulate::use_virtualenv("/opt/venv/scanorama")
reticulate::py_discover_config()
```

```{r}
# download pre-computed data if missing or long compute
fetch_data <- TRUE

# url for source and intermediate data
path_data <- "https://export.uppmax.uu.se/naiss2023-23-3/workshops/workshop-scrnaseq"
path_file <- "data/covid/results/seurat_covid_qc_dr.rds"
if (!dir.exists(dirname(path_file))) dir.create(dirname(path_file), recursive = TRUE)
if (fetch_data && !file.exists(path_file)) download.file(url = file.path(path_data, "covid/results/seurat_covid_qc_dr.rds"), destfile = path_file)
alldata <- readRDS(path_file)
print(names(alldata@reductions))
```

{{< meta int_prep_2 >}}

```{r}
#| fig-height: 6
#| fig-width: 8

alldata.list <- SplitObject(alldata, split.by = "orig.ident")

for (i in 1:length(alldata.list)) {
    alldata.list[[i]] <- NormalizeData(alldata.list[[i]], verbose = FALSE)
    alldata.list[[i]] <- FindVariableFeatures(alldata.list[[i]], selection.method = "vst", nfeatures = 2000,verbose = FALSE)
}

# get the variable genes from all the datasets.
hvgs_per_dataset <- lapply(alldata.list, function(x) { x@assays$RNA@var.features })

# also add in the variable genes that was selected on the whole dataset
hvgs_per_dataset$all = VariableFeatures(alldata)

temp <- unique(unlist(hvgs_per_dataset))
overlap <- sapply( hvgs_per_dataset , function(x) { temp %in% x } )
pheatmap::pheatmap(t(overlap*1),cluster_rows = F ,
                   color = c("grey90","grey20"))
```
As you can see, there are a lot of genes that are variable in just one dataset. There are also some genes in the gene set that was selected using all the data that are not variable in any of the individual datasets. These are most likely genes driven by batch effects. 

A better way to select features for integration is to combine the information on variable genes across the dataset. This can be done with the function `SelectIntegrationFeatures` that combines the ranks of the variable features in the different datasets.

```{r}
hvgs_all = SelectIntegrationFeatures(alldata.list)
hvgs_per_dataset$all_ranks = hvgs_all

temp <- unique(unlist(hvgs_per_dataset))
overlap <- sapply( hvgs_per_dataset , function(x) { temp %in% x } )
pheatmap::pheatmap(t(overlap*1),cluster_rows = F ,
                   color = c("grey90","grey20"))
```

For all downstream integration we will use this set of genes. 

## CCA

We identify anchors using the `FindIntegrationAnchors()` function, which takes a list of Seurat objects as input.

```{r}
alldata.anchors <- FindIntegrationAnchors(object.list = alldata.list, dims = 1:30,reduction = "cca", anchor.features = hvgs_all)
```

We then pass these anchors to the `IntegrateData()` function, which returns a Seurat object.

```{r}
alldata.int <- IntegrateData(anchorset = alldata.anchors, dims = 1:30, new.assay.name = "CCA")
```

We can observe that a new assay slot is now created under the name `CCA`. If you do not specify the assay name, the default will be `integrated`. 

```{r}
names(alldata.int@assays)

# by default, Seurat now sets the integrated assay as the default assay, so any operation you now perform will be on the integrated data.
alldata.int@active.assay
```

After running `IntegrateData()`, the Seurat object will contain a new Assay with the integrated (or **batch-corrected**) expression matrix. Note that the original (uncorrected values) are still stored in the object in the “RNA” assay, so you can switch back and forth. We can then use this new integrated matrix for downstream analysis and visualization. Here we scale the integrated data, run PCA, and visualize the results with UMAP and TSNE. The integrated datasets cluster by cell type, instead of by technology.

As `CCA` is the `active.assay` now, the functions will by default run on the data in that assay. But you could also specify in each of the functions to run them in a specific assay with the parameter `assay = "CCA"`.

```{r}
#Run Dimensionality reduction on integrated space
alldata.int <- ScaleData(alldata.int, verbose = FALSE)
alldata.int <- RunPCA(alldata.int, npcs = 30, verbose = FALSE)
alldata.int <- RunUMAP(alldata.int, dims = 1:30)
alldata.int <- RunTSNE(alldata.int, dims = 1:30)
```

{{< meta int_plot >}}

```{r}
#| fig-height: 6
#| fig-width: 10

wrap_plots(
  DimPlot(alldata, reduction = "pca", group.by = "orig.ident")+NoAxes()+ggtitle("PCA raw_data"),
  DimPlot(alldata, reduction = "tsne", group.by = "orig.ident")+NoAxes()+ggtitle("tSNE raw_data"),
  DimPlot(alldata, reduction = "umap", group.by = "orig.ident")+NoAxes()+ggtitle("UMAP raw_data"),
  
  DimPlot(alldata.int, reduction = "pca", group.by = "orig.ident")+NoAxes()+ggtitle("PCA integrated"),
  DimPlot(alldata.int, reduction = "tsne", group.by = "orig.ident")+NoAxes()+ggtitle("tSNE integrated"),
  DimPlot(alldata.int, reduction = "umap", group.by = "orig.ident")+NoAxes()+ggtitle("UMAP integrated"),
  ncol = 3
) + plot_layout(guides = "collect")
```

### Clean memory

Again we have a lot of large objects in the memory. We have the original data `alldata` but also the integrated data in `alldata.int`.  We also have the split objects in `alldata.list` and the anchors in `alldata.anchors`. In principle we only need the integrated object for now, but we will also keep the list for running Scanorama further down in the tutorial.

We also want to keep the original umap for visualization purposes, so we copy it over to the integrated object.

```{r}
alldata.int@reductions$umap_raw = alldata@reductions$umap

# remove all objects that will not be used.
rm(alldata,  alldata.anchors)
# run garbage collect to free up memory
gc()
```


{{< meta dimred_plotgenes_1 >}}

|Markers|Cell Type|
|:---|:---|
|CD3E|T cells|
|CD3E CD4|CD4+ T cells|
|CD3E CD8A|CD8+ T cells|
|GNLY, NKG7|NK cells|
|MS4A1|B cells|
|CD14, LYZ, CST3, MS4A7|CD14+ Monocytes|
|FCGR3A, LYZ, CST3, MS4A7|FCGR3A+  Monocytes|
|FCER1A, CST3|DCs|

```{r}
#| fig-height: 8
#| fig-width: 10

myfeatures <- c("CD3E", "CD4", "CD8A", "NKG7", "GNLY", "MS4A1", "CD14", "LYZ", "MS4A7", "FCGR3A", "CST3", "FCER1A")
FeaturePlot(alldata.int, reduction = "umap", dims = 1:2, features = myfeatures, ncol = 4, order = T) + NoLegend() + NoAxes() + NoGrid()
```


## {{< meta dimred_harmony >}}

{{< meta dimred_harmony_1 >}}

OBS! Make sure to revert back to the `RNA` assay.

```{r}
alldata.int@active.assay = "RNA"
VariableFeatures(alldata.int) = hvgs_all
alldata.int = ScaleData(alldata.int, vars.to.regress = c("percent_mito", "nFeature_RNA"))
alldata.int = RunPCA(alldata.int, reduction.name = "pca_harmony")
```

Now we are ready to run Harmony.

```{r}
#| fig-height: 10
#| fig-width: 13

library(harmony)

alldata.int <- RunHarmony(
  alldata.int,
  group.by.vars = "orig.ident",
  reduction.use = "pca_harmony",
  dims.use = 1:50,
  assay.use = "RNA")
```

Harmony will create another reduction slot in your seurat object with the name **harmony**, so now we can use that reduction instead of PCA to run UMAP.

```{r}
alldata.int <- RunUMAP(alldata.int, dims = 1:50, reduction = "harmony", reduction.name = "umap_harmony")

DimPlot(alldata.int, reduction = "umap_harmony", group.by = "orig.ident") + NoAxes() + ggtitle("Harmony UMAP")
```

## {{< meta dimred_scanorama >}}

:::{.callout-important}
If you are running locally using Docker and you have a Mac with ARM chip, the Scanorama reticulate module is known to crash. In this case, you might want to skip this section.
:::

{{< meta dimred_scanorama_1 >}}

{{< meta dimred_scanorama_2 >}}


```{r}
#| fig-height: 5
#| fig-width: 16

assaylist <- list()
genelist <- list()
for(i in 1:length(alldata.list)) {
  assaylist[[i]] <- t(as.matrix(GetAssayData(alldata.list[[i]], "data")[hvgs_all,]))
  genelist[[i]] <- hvgs_all
}

lapply(assaylist,dim)
```

Then, we use the `scanorama` function through reticulate. The integrated data is added back into the Seurat object as a new Reduction. 

```{r}
#| fig-height: 5
#| fig-width: 16

# Activate scanorama Python venv
scanorama <- reticulate::import("scanorama")

integrated.data <- scanorama$integrate(datasets_full = assaylist,
                                       genes_list = genelist )

# Now we create a new dim reduction object in the format that Seurat uses
intdimred <- do.call(rbind, integrated.data[[1]])
colnames(intdimred) <- paste0("PC_", 1:100)
rownames(intdimred) <- colnames(alldata.int)

# Add standard deviations in order to draw Elbow Plots in Seurat
stdevs <- apply(intdimred, MARGIN = 2, FUN = sd)

# Create a new dim red object.
alldata.int[["scanorama"]] <- CreateDimReducObject(
  embeddings = intdimred,
  stdev      = stdevs,
  key        = "PC_",
  assay      = "RNA")
```

```{r}
#Here we use all PCs computed from Scanorama for UMAP calculation
alldata.int <- RunUMAP(alldata.int, dims = 1:100, reduction = "scanorama",reduction.name = "umap_scanorama")

DimPlot(alldata.int, reduction = "umap_scanorama", group.by = "orig.ident") + NoAxes() + ggtitle("Harmony UMAP")
```

## Overview all methods

Now we will plot UMAPS with all three integration methods side by side. 

```{r}
#| fig-height: 8
#| fig-width: 9

p1 <- DimPlot(alldata.int, reduction = "umap_raw", group.by = "orig.ident") + ggtitle("UMAP raw_data")
p2 <- DimPlot(alldata.int, reduction = "umap", group.by = "orig.ident") + ggtitle("UMAP CCA")
p3 <- DimPlot(alldata.int, reduction = "umap_harmony", group.by = "orig.ident") + ggtitle("UMAP Harmony")
p4 <- DimPlot(alldata.int, reduction = "umap_scanorama", group.by = "orig.ident")+ggtitle("UMAP Scanorama")

wrap_plots(p1, p2, p3, p4, nrow = 2) + plot_layout(guides = "collect")
```


::: {.callout-note title="Discuss"}
{{< meta int_conclusion >}}
:::


{{< meta int_save >}}

```{r}
saveRDS(alldata.int,"data/covid/results/seurat_covid_qc_dr_int.rds")
```


## Extra task

You have now done the Seurat integration with CCA which is quite slow. There are other options in the `FindIntegrationAnchors()` function. Try rerunning the integration with `rpca` and/or `rlsi` and create a new UMAP. Compare the results.

## {{< meta session >}}

<details>
  <summary>Click here</summary>

```{r}
sessionInfo()
```

</details>



