---
title: "{{< meta dimred_title >}}"
subtitle: "{{< meta subtitle_seurat >}}"
description: "{{< meta dimred_description >}}"
format: html
---

::: {.callout-note}
Code chunks run R commands unless otherwise specified.
:::

## {{< meta dimred_prep >}}

{{< meta dimred_prep_1 >}}

```{r}
#| label: libraries

suppressPackageStartupMessages({
    library(Seurat)
    library(ggplot2) # plotting
    library(patchwork) # combining figures
    library(scran)
})
```

```{r}
#| label: fetch-data

# download pre-computed data if missing or long compute
fetch_data <- TRUE

# url for source and intermediate data
path_data <- "https://nextcloud.dc.scilifelab.se/public.php/webdav"
curl_upass <- "-u zbC5fr2LbEZ9rSE:scRNAseq2025"

path_file <- "data/covid/results/seurat_covid_qc.rds"
if (!dir.exists(dirname(path_file))) dir.create(dirname(path_file), recursive = TRUE)
if (fetch_data && !file.exists(path_file)) download.file(url = file.path(path_data, "covid/results_seurat/seurat_covid_qc.rds"), destfile = path_file, method = "curl", extra = curl_upass)

alldata <- readRDS(path_file)
```

## {{< meta dimred_fs >}}

{{< meta dimred_fs_2 >}}

```{r}
#| label: hvg

suppressWarnings(suppressMessages(alldata <- FindVariableFeatures(alldata, selection.method = "vst", nfeatures = 2000, verbose = FALSE, assay = "RNA")))
top20 <- head(VariableFeatures(alldata), 20)

LabelPoints(plot = VariableFeaturePlot(alldata), points = top20, repel = TRUE)
```

## {{< meta dimred_zs >}}

{{< meta dimred_zs_1 >}}

```{r}
#| label: scale

alldata <- ScaleData(alldata, vars.to.regress = c("percent_mito", "nFeature_RNA"), assay = "RNA")
```

## {{< meta dimred_pca >}}

{{< meta dimred_pca_1 >}}

To run PCA, you can use the function `RunPCA()`.

```{r}
#| label: pca

alldata <- RunPCA(alldata, npcs = 50, verbose = F)
```

{{< meta dimred_pca_2 >}}

```{r}
#| label: pca-plot
#| fig-height: 4
#| fig-width: 12

wrap_plots(
    DimPlot(alldata, reduction = "pca", group.by = "orig.ident", dims = 1:2),
    DimPlot(alldata, reduction = "pca", group.by = "orig.ident", dims = 3:4),
    DimPlot(alldata, reduction = "pca", group.by = "orig.ident", dims = 5:6),
    ncol = 3
) + plot_layout(guides = "collect")
```

{{< meta dimred_pca_3 >}}

```{r}
#| label: pca-loadings
#| fig-height: 6
#| fig-width: 14

VizDimLoadings(alldata, dims = 1:5, reduction = "pca", ncol = 5, balanced = T)
```

{{< meta dimred_pca_4 >}}

```{r}
#| label: pca-elbow
#| fig-height: 4
#| fig-width: 5

ElbowPlot(alldata, reduction = "pca", ndims = 50)
```

{{< meta dimred_pca_5 >}}

With the `scater` package we can check how different metadata variables contribute to each PCs.
This can be important to look at to understand different biases you may have in your data.


```{r}
#| label: pca-explanatory
#| fig-height: 4
#| fig-width: 10

scater::plotExplanatoryPCs(as.SingleCellExperiment(alldata), nvars_to_plot = 15, npcs_to_plot = 20)
```

:::{.callout-note title="Discuss"}
{{< meta dimred_pca_6 >}}
:::

## {{< meta dimred_tsne >}}

{{< meta dimred_tsne_1 >}}

```{r}
#| label: run-tsne

alldata <- RunTSNE(
    alldata,
    reduction = "pca", dims = 1:30,
    perplexity = 30,
    max_iter = 1000,
    theta = 0.5,
    eta = 200,
    num_threads = 0
)
# see ?Rtsne and ?RunTSNE for more info
```

{{< meta dimred_tsne_2 >}}

```{r}
#| label: plot-tsne
#| fig-height: 5
#| fig-width: 6

DimPlot(alldata, reduction = "tsne", group.by = "orig.ident")
```


## {{< meta dimred_umap >}}

{{< meta dimred_umap_1 >}}

```{r}
#| label: run-umap

alldata <- RunUMAP(
    alldata,
    reduction = "pca",
    dims = 1:30,
    n.components = 2,
    n.neighbors = 30,
    n.epochs = 200,
    min.dist = 0.3,
    learning.rate = 1,
    spread = 1
)
# see ?RunUMAP for more info
```

A feature of UMAP is that it is not limited by the number of dimensions the data cen be reduced into (unlike tSNE). We can simply reduce the dimentions altering the `n.components` parameter. So here we will create a UMAP with 10 dimensions.

In Seurat, we can add in additional reductions, by default they are named "pca", "umap", "tsne" etc. depending on the function you run. Here we will specify an alternative name for the umap with the `reduction.name` parameter. 

```{r}
#| label: run-umap2

alldata <- RunUMAP(
    alldata,
    reduction.name = "UMAP10_on_PCA",
    reduction = "pca",
    dims = 1:30,
    n.components = 10,
    n.neighbors = 30,
    n.epochs = 200,
    min.dist = 0.3,
    learning.rate = 1,
    spread = 1
)
# see ?RunUMAP for more info
```

{{< meta dimred_umap_2 >}}

```{r}
#| label: plot-umap
#| fig-height: 4
#| fig-width: 12

wrap_plots(
    DimPlot(alldata, reduction = "umap", group.by = "orig.ident") + ggplot2::ggtitle(label = "UMAP_on_PCA"),
    DimPlot(alldata, reduction = "UMAP10_on_PCA", group.by = "orig.ident", dims = 1:2) + ggplot2::ggtitle(label = "UMAP10_on_PCA"),
    DimPlot(alldata, reduction = "UMAP10_on_PCA", group.by = "orig.ident", dims = 3:4) + ggplot2::ggtitle(label = "UMAP10_on_PCA"),
    ncol = 3
) + plot_layout(guides = "collect")
```

{{< meta dimred_umap_3 >}}

```{r}
#| label: plot-dimred
#| fig-height: 4
#| fig-width: 12

wrap_plots(
    DimPlot(alldata, reduction = "pca", group.by = "orig.ident"),
    DimPlot(alldata, reduction = "tsne", group.by = "orig.ident"),
    DimPlot(alldata, reduction = "umap", group.by = "orig.ident"),
    ncol = 3
) + plot_layout(guides = "collect")
```

::: {.callout-note title="Discuss"}
{{< meta dimred_umap_4 >}}
:::

## {{< meta dimred_zsg >}}

{{< meta dimred_zsg_1 >}}

### {{< meta dimred_zsg_zs >}}

{{< meta dimred_zsg_zs_1 >}}

```{r}
#| label: run-umap-sd

alldata <- RunUMAP(
    alldata,
    reduction.name = "UMAP_on_ScaleData",
    features = VariableFeatures(alldata),
    assay = "RNA",
    n.components = 2,
    n.neighbors = 30,
    n.epochs = 200,
    min.dist = 0.3,
    learning.rate = 1,
    spread = 1
)
```

### {{< meta dimred_zsg_g >}}

{{< meta dimred_zsg_g_1 >}}

```{r}
#| label: umap-graph
#| eval: false

#OBS! Skip for now, known issue with later version of umap-learn in Seurat5
# have 0.5.7 now, tested downgrading to 0.5.4 or 0.5.3 but still have same error.
# Seurat 5.2.0 has a fix for this, but not the version we have now.

# Build Graph
alldata <- FindNeighbors(alldata,
    reduction = "pca",
    assay = "RNA",
    k.param = 20,
    features = VariableFeatures(alldata)
)

alldata <- RunUMAP(alldata,
    reduction.name = "UMAP_on_Graph",
    umap.method = "umap-learn",
    graph = "RNA_snn",
    n.epochs = 200,
    assay = "RNA"
)
```

{{< meta dimred_zsg_g_2 >}}

```{r}
#| label: plot-umap-graph
#| eval: true
#| fig-height: 4
#| fig-width: 12

p1 <- DimPlot(alldata, reduction = "umap", group.by = "orig.ident") + ggplot2::ggtitle(label = "UMAP_on_PCA")
p2 <- DimPlot(alldata, reduction = "UMAP_on_ScaleData", group.by = "orig.ident") + ggplot2::ggtitle(label = "UMAP_on_ScaleData")
#p3 <- DimPlot(alldata, reduction = "UMAP_on_Graph", group.by = "orig.ident") + ggplot2::ggtitle(label = "UMAP_on_Graph")
#wrap_plots(p1, p2, p3, ncol = 3) + plot_layout(guides = "collect")
wrap_plots(p1, p2, ncol = 2) + plot_layout(guides = "collect")
```

## {{< meta dimred_plotgenes >}}

{{< meta dimred_plotgenes_1 >}}

|Markers|Cell Type|
|:---|:---|
|CD3E|T cells|
|CD3E CD4|CD4+ T cells|
|CD3E CD8A|CD8+ T cells|
|GNLY, NKG7|NK cells|
|MS4A1|B cells|
|CD14, LYZ, CST3, MS4A7|CD14+ Monocytes|
|FCGR3A, LYZ, CST3, MS4A7|FCGR3A+  Monocytes|
|FCER1A, CST3|DCs|

```{r}
#| label: plot-markers
#| fig-height: 9
#| fig-width: 12

myfeatures <- c("CD3E", "CD4", "CD8A", "NKG7", "GNLY", "MS4A1", "CD14", "LYZ", "MS4A7", "FCGR3A", "CST3", "FCER1A")
FeaturePlot(alldata, reduction = "umap", dims = 1:2, features = myfeatures, ncol = 4, order = T)
```

:::{.callout-note title="Discuss"}
{{< meta dimred_plotgenes_2 >}}
:::

```{r}
#| label: plot-qc
#| fig-height: 9
#| fig-width: 12

myfeatures <- c("nCount_RNA","nFeature_RNA", "percent_mito","percent_ribo","percent_hb","percent_plat")
FeaturePlot(alldata, reduction = "umap", dims = 1:2, features = myfeatures, ncol = 3, order = T) 
```


## {{< meta dimred_save >}}

{{< meta dimred_save_1 >}}

```{r}
#| label: save

saveRDS(alldata, "data/covid/results/seurat_covid_qc_dr.rds")
```

## {{< meta session >}}

<details>
  <summary>Click here</summary>

```{r}
#| label: session

sessionInfo()
```

</details>
