---
title: "scran_analysis.Rmd"
output:
  md_document:
    variant: markdown_github
---

# Analysis with the scran package

Analysis of data with scran/scater package, using workflow at:
https://www.bioconductor.org/help/workflows/simpleSingleCell/

Written by Asa K. Bjorklund, 2017

### Load packages
```{r}
suppressMessages(require(scran))
suppressMessages(require(scater))
```

### Load expression values and metadata

```{r}
D <- read.table("data/ILC/ensembl_countvalues_ILC.csv",sep=",",header=T,row.names=1)
M <- read.table("data/ILC/Metadata_ILC.csv",sep=",",header=T,row.names=1)
pd <- new("AnnotatedDataFrame", data=M)

# check which genes are spike-ins in the matrix
spikes <- grepl("ERCC_",rownames(D))

```

## Calculate qc-metrics

Using scran package

```{r}
sce <- newSCESet(countData=D,phenoData=pd)
dim(sce)
sce <- calculateQCMetrics(sce, feature_controls=list(ERCC=spikes))
setSpike(sce) <- "ERCC"

# plot library size and detected genes
par(mfrow=c(1,2))
hist(sce$total_counts/1e6, xlab="Library sizes (millions)", main="", 
    breaks=20, col="grey80", ylab="Number of cells")
hist(sce$total_features, xlab="Number of expressed genes", main="", 
    breaks=20, col="grey80", ylab="Number of cells")

```

### PCA plot

```{r}
fontsize <- theme(axis.text=element_text(size=12), axis.title=element_text(size=16))
plotPCA(sce, pca_data_input="pdata",colour_by="Plate") + fontsize

```
We have some outlier cells that clearly deviates

### Cell cycle prediction 

To use the predefined sets of cell cycle markers that they have prepared for mouse and human. You need to have Ensembl gene IDs for your genes. If you do not have them, take a look at the [biomaRt tutorial](biomaRt.md).

```{r}

hsa.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", package="scran"))
assignments <- cyclone(sce, hsa.pairs)
plot(assignments$score$G1, assignments$score$G2M, xlab="G1 score", ylab="G2/M score", pch=16)

head(assignments$scores)
# check predictions
table(assignments$phases)

# plot pca and color by cell cycle predictions, check if clear separation by CC
# add phases to phenoData
phenoData(sce)$cc.phases <- assignments$phases
plotPCA(sce, pca_data_input="exprs",colour_by="cc.phases",ncomponents=5) + fontsize
# find PCs that correlates with predicted phase
plotQC(sce, type = "find-pcs", variable = "cc.phases",plot_type = "pairs-pcs") + fontsize
```

Does not seem to be a clear separation by cell-cycle phase for this dataset.

### Filter out low abundance genes

```{r}
ave.counts <- calcAverage(sce)
keep <- ave.counts >= 1
sum(keep)

# check distribution of average counts and check if cutoff makes sense.
hist(log10(ave.counts), breaks=100, main="", col="grey80",
    xlab=expression(Log[10]~"average count"))
abline(v=log10(1), col="blue", lwd=2, lty=2)

# plot the most highly expressed genes.
plotQC(sce, type = "highest-expression", n=50) + fontsize

# calculate number of cells that each gene is found in
numcells <- nexprs(sce, byrow=TRUE)

#The relationship between the number of expressing cells and the mean. The two statistics tend to be well-correlated so filtering on either should give roughly similar results.
smoothScatter(log10(ave.counts), numcells, xlab=expression(Log[10]~"average count"), 
    ylab="Number of expressing cells")
is.ercc <- isSpike(sce, type="ERCC")
points(log10(ave.counts[is.ercc]), numcells[is.ercc], col="red", pch=16, cex=0.5)


```

### Normalization 

```{r}
sce <- computeSumFactors(sce, sizes=seq(20, 80, 5))
summary(sizeFactors(sce))

high.ab <- calcAverage(sce) > 1
sce <- computeSumFactors(sce, subset.row=high.ab)
summary(sizeFactors(sce))
```

