---
title: "{{< meta int_title >}}"
subtitle: "{{< meta subtitle_seurat >}}"
description: "{{< meta int_description >}}"
format: html
---

::: {.callout-note}
Code chunks run R commands unless otherwise specified.
:::

{{< meta int_1 >}}

|Markdown | Language | Library | Ref|
|:---|:---|:---|:---|
|CCA | R | Seurat | [Cell](https://www.sciencedirect.com/science/article/pii/S0092867419305598?via%3Dihub)|
|MNN | R/Python | Scater/Scanpy | [Nat. Biotech.](https://www.nature.com/articles/nbt.4091)|
|Conos | R | conos | [Nat. Methods](https://www.nature.com/articles/s41592-019-0466-z?error=cookies_not_supported&code=5680289b-6edb-40ad-9934-415dac4fdb2f)|
|Harmony | R | Harmony | [Nat. Methods](https://www.nature.com/articles/s41592-019-0619-0)|
|Scanorama | Python | scanorama | [Nat. Biotech.](https://www.nature.com/articles/s41587-019-0113-3)|


## {{< meta int_prep >}}

{{< meta int_prep_1 >}}

```{r}
#| label: libraries
suppressPackageStartupMessages({
    library(Seurat)
    library(ggplot2)
    library(patchwork)
    library(basilisk)
})

condapath = "/usr/local/conda/envs/seurat"
```

```{r}
#| label: fetch-data

# download pre-computed data if missing or long compute
fetch_data <- TRUE

# url for source and intermediate data
path_data <- "https://nextcloud.dc.scilifelab.se/public.php/webdav"
curl_upass <- "-u zbC5fr2LbEZ9rSE:scRNAseq2025"

path_file <- "data/covid/results/seurat_covid_qc_dr.rds"
if (!dir.exists(dirname(path_file))) dir.create(dirname(path_file), recursive = TRUE)
if (fetch_data && !file.exists(path_file)) download.file(url = file.path(path_data, "covid/results_seurat/seurat_covid_qc_dr.rds"), destfile = path_file, method = "curl", extra = curl_upass)

alldata <- readRDS(path_file)
print(names(alldata@reductions))
```

With Seurat5 we can split the `RNA` assay into multiple `Layers` with one count matrix and one data matrix per sample. When we then run `FindVariableFeatures` on the object it will run it for each of the samples separately, but also compute the overall variable features by combining their ranks.

```{r}
#| label: hvg
#| fig-height: 6
#| fig-width: 8

# get the variable genes from all the datasets without batch information.
hvgs_old = VariableFeatures(alldata)

# now split the object into layers
alldata[["RNA"]] <- split(alldata[["RNA"]], f = alldata$orig.ident)

# detect HVGs
alldata <- FindVariableFeatures(alldata, selection.method = "vst", nfeatures = 2000, verbose = FALSE)

# to get the HVGs for each layer we have to fetch them individually
data.layers <- Layers(alldata)[grep("data.",Layers(alldata))]
print(data.layers)

hvgs_per_dataset <- lapply(data.layers, function(x) VariableFeatures(alldata, layer = x) )
names(hvgs_per_dataset) = data.layers

# also add in the variable genes that was selected on the whole dataset and the old ones 
hvgs_per_dataset$all <- VariableFeatures(alldata)
hvgs_per_dataset$old <- hvgs_old

temp <- unique(unlist(hvgs_per_dataset))
overlap <- sapply( hvgs_per_dataset , function(x) { temp %in% x } )
pheatmap::pheatmap(t(overlap*1),cluster_rows = F ,
                   color = c("grey90","grey20"))
```
As you can see, there are a lot of genes that are variable in just one dataset. There are also some genes in the gene set that was selected using all the data that are not variable in any of the individual datasets. These are most likely genes driven by batch effects. 

A better way to select features for integration is to combine the information on variable genes across the dataset. This is what we have in the `all` section where the ranks of the variable features in the different datasets is combined.

:::{.callout-note title="Discuss"}
{{< meta int_hvg >}}
:::

For all downstream integration we will use this set of genes so that it is comparable across the methods. Before doing anything else we need to rerun `ScaleData` and `PCA` with that set of genes.


```{r}
#| label: pca
hvgs_all = hvgs_per_dataset$all

alldata = ScaleData(alldata, features = hvgs_all, vars.to.regress = c("percent_mito", "nFeature_RNA"))
alldata = RunPCA(alldata, features = hvgs_all, verbose = FALSE)
```

## CCA

In Seurat v4 we run the integration in two steps, first finding anchors between datasets with `FindIntegrationAnchors()` and then running the actual integration with `IntegrateData()`. Since Seurat v5 this is done in a single command using the function `IntegrateLayers()`, we specify the name for the integration as `integrated_cca`.

``` {r}
#| label: run-cca
alldata <- IntegrateLayers(object = alldata, 
                           method = CCAIntegration, orig.reduction = "pca", 
                           new.reduction = "integrated_cca", verbose = FALSE)
```

We should now have a new dimensionality reduction slot (`integrated_cca`) in the object:

```{r}
#| label: reductions
names(alldata@reductions)
```

Using this new integrated dimensionality reduction we can now run UMAP and tSNE on that object, and we again specify the names of the new reductions so that the old UMAP and tSNE are not overwritten.  

``` {r}
#| label: proc-cca
alldata <- RunUMAP(alldata, reduction = "integrated_cca", dims = 1:30, reduction.name = "umap_cca")
alldata <- RunTSNE(alldata, reduction = "integrated_cca", dims = 1:30, reduction.name = "tsne_cca")

names(alldata@reductions)
```

{{< meta int_plot >}}

```{r}
#| label: plot-cca
#| fig-height: 6
#| fig-width: 10

wrap_plots(
  DimPlot(alldata, reduction = "pca", group.by = "orig.ident")+NoAxes()+ggtitle("PCA raw_data"),
  DimPlot(alldata, reduction = "tsne", group.by = "orig.ident")+NoAxes()+ggtitle("tSNE raw_data"),
  DimPlot(alldata, reduction = "umap", group.by = "orig.ident")+NoAxes()+ggtitle("UMAP raw_data"),
  
  DimPlot(alldata, reduction = "integrated_cca", group.by = "orig.ident")+NoAxes()+ggtitle("CCA integrated"),
  DimPlot(alldata, reduction = "tsne_cca", group.by = "orig.ident")+NoAxes()+ggtitle("tSNE integrated"),
  DimPlot(alldata, reduction = "umap_cca", group.by = "orig.ident")+NoAxes()+ggtitle("UMAP integrated"),
  ncol = 3
) + plot_layout(guides = "collect")

```

### Marker genes

{{< meta dimred_plotgenes_1 >}}

|Markers|Cell Type|
|:---|:---|
|CD3E|T cells|
|CD3E CD4|CD4+ T cells|
|CD3E CD8A|CD8+ T cells|
|GNLY, NKG7|NK cells|
|MS4A1|B cells|
|CD14, LYZ, CST3, MS4A7|CD14+ Monocytes|
|FCGR3A, LYZ, CST3, MS4A7|FCGR3A+  Monocytes|
|FCER1A, CST3|DCs|

```{r}
#| label: plot-markers
#| fig-height: 8
#| fig-width: 10

myfeatures <- c("CD3E", "CD4", "CD8A", "NKG7", "GNLY", "MS4A1", "CD14", "LYZ", "MS4A7", "FCGR3A", "CST3", "FCER1A")
FeaturePlot(alldata, reduction = "umap_cca", dims = 1:2, features = myfeatures, ncol = 4, order = T) + NoLegend() + NoAxes() + NoGrid()
```

## {{< meta dimred_harmony >}}

{{< meta dimred_harmony_1 >}}

We can use the same function `IntegrateLayers()` but intstead specify the method `HarmonyIntegration`. And as above, we run UMAP on the new reduction from Harmony.

```{r}
#| label: harmony
#| fig-height: 10
#| fig-width: 13

alldata <- IntegrateLayers(
  object = alldata, method = HarmonyIntegration,
  orig.reduction = "pca", new.reduction = "harmony",
  verbose = FALSE
)


alldata <- RunUMAP(alldata, dims = 1:30, reduction = "harmony", reduction.name = "umap_harmony")
DimPlot(alldata, reduction = "umap_harmony", group.by = "orig.ident") + NoAxes() + ggtitle("Harmony UMAP")
```

## {{< meta dimred_scanorama >}}

{{< meta dimred_scanorama_1 >}}

{{< meta dimred_scanorama_2 >}}

Keep in mind that for most python tools that uses `AnnData` format the gene x cell matrix is transposed
so that genes are rows and cells are columns. 

```{r}
#| label: prep-scanorama
# get data matrices from all samples, with only the variable genes.
data.layers <- Layers(alldata)[grep("data.",Layers(alldata))]
print(data.layers)

assaylist <- lapply(data.layers, function(x) t(as.matrix(LayerData(alldata, layer = x)[hvgs_all,])))
genelist =  rep(list(hvgs_all),length(assaylist))

lapply(assaylist,dim)
```

Scanorama is implemented in python, but through reticulate we can load python packages and run python functions. In this case we also use the `basilisk` package for a more clean activation of python environment.

At the top of this script, we set the variable `condapath` to point to the conda environment where scanorama is included. 

```{r}
#| label: prep-scanorama2

# run scanorama via basilisk with assaylis and genelist as input.
integrated.data = basiliskRun(env=condapath, fun=function(datas, genes) {
  scanorama <- reticulate::import("scanorama")
  output <- scanorama$integrate(datasets_full = datas,
                                         genes_list = genes )
  return(output)
}, datas = assaylist, genes = genelist, testload="scanorama")


# Now we create a new dim reduction object in the format that Seurat uses
intdimred <- do.call(rbind, integrated.data[[1]])
colnames(intdimred) <- paste0("Scanorama_", 1:100)
rownames(intdimred) <- colnames(alldata)

# Add standard deviations in order to draw Elbow Plots in Seurat
stdevs <- apply(intdimred, MARGIN = 2, FUN = sd)

# Create a new dim red object.
alldata[["scanorama"]] <- CreateDimReducObject(
  embeddings = intdimred,
  stdev      = stdevs,
  key        = "Scanorama_",
  assay      = "RNA")
```

Try the same but using counts instead of data.

``` {r}
#| label: run-scanorama
# get count matrices from all samples, with only the variable genes.
count.layers <- Layers(alldata)[grep("counts.",Layers(alldata))]
print(count.layers)
assaylist <- lapply(count.layers, function(x) t(as.matrix(LayerData(alldata, layer = x)[hvgs_all,])))

# run scanorama	via basilisk with assaylis and genelist	as input.
integrated.data = basiliskRun(env=condapath, fun=function(datas, genes) {
  scanorama <- reticulate::import("scanorama")
  output <- scanorama$integrate(datasets_full = datas,
                                         genes_list = genes )
  return(output)
}, datas = assaylist, genes = genelist, testload="scanorama")

# Now we create a new dim reduction object in the format that Seurat uses
# The scanorama output has 100 dimensions.
intdimred <- do.call(rbind, integrated.data[[1]])
colnames(intdimred) <- paste0("Scanorama_", 1:100)
rownames(intdimred) <- colnames(alldata)

# Add standard deviations in order to draw Elbow Plots in Seurat
stdevs <- apply(intdimred, MARGIN = 2, FUN = sd)

# Create a new dim red object.
alldata[["scanoramaC"]] <- CreateDimReducObject(
  embeddings = intdimred,
  stdev      = stdevs,
  key        = "Scanorama_",
  assay      = "RNA")
```

```{r}
#| label: plot-scanorama
#| fig-height: 8
#| fig-width: 14

#Here we use all PCs computed from Scanorama for UMAP calculation
alldata <- RunUMAP(alldata, dims = 1:100, reduction = "scanorama",reduction.name = "umap_scanorama")
alldata <- RunUMAP(alldata, dims = 1:100, reduction = "scanoramaC",reduction.name = "umap_scanoramaC")


p1 = DimPlot(alldata, reduction = "umap_scanorama", group.by = "orig.ident") + NoAxes() + ggtitle("Scanorama UMAP")
p2 = DimPlot(alldata, reduction = "umap_scanoramaC", group.by = "orig.ident") + NoAxes() + ggtitle("ScanoramaC UMAP")

wrap_plots(p1,p2)
```

## Overview all methods

Now we will plot UMAPS with all three integration methods side by side. 

```{r}
#| label: plot-all
#| fig-height: 8
#| fig-width: 9

p1 <- DimPlot(alldata, reduction = "umap", group.by = "orig.ident") + ggtitle("UMAP raw_data")
p2 <- DimPlot(alldata, reduction = "umap_cca", group.by = "orig.ident") + ggtitle("UMAP CCA")
p3 <- DimPlot(alldata, reduction = "umap_harmony", group.by = "orig.ident") + ggtitle("UMAP Harmony")
p4 <- DimPlot(alldata, reduction = "umap_scanorama", group.by = "orig.ident")+ggtitle("UMAP Scanorama")

wrap_plots(p1, p2, p3, p4, nrow = 2) + plot_layout(guides = "collect")
```


::: {.callout-note title="Discuss"}
{{< meta int_conclusion >}}
:::


{{< meta int_save >}}

```{r}
#| label: save

saveRDS(alldata,"data/covid/results/seurat_covid_qc_dr_int.rds")
```


## Extra task

You have now done the Seurat integration with CCA which is quite slow. There are other options in the `IntegrateLayers()` function. Try rerunning the integration with `RPCAIntegration` and create a new UMAP. Compare the results.

## {{< meta session >}}

<details>
  <summary>Click here</summary>

```{r}
#| label: session

sessionInfo()
```

</details>



