---
title: "{{< meta ct_title >}}"
subtitle: "{{< meta subtitle_bioc >}}"
description: "{{< meta ct_description >}}"
format: html
---

::: {.callout-note}
Code chunks run R commands unless otherwise specified.
:::

{{< meta ct_1 >}}

{{< meta ct_2 >}}

## {{< meta ct_read >}}

{{< meta ct_read_1 >}}

```{r}
suppressPackageStartupMessages({
    library(scater)
    library(scran)
    library(dplyr)
    library(patchwork)
    library(ggplot2)
    library(pheatmap)
    library(scPred)
    library(scmap)
})
```

{{< meta ct_read_2 >}}

```{r}
# download pre-computed data if missing or long compute
fetch_data <- TRUE

# url for source and intermediate data
path_data <- "https://export.uppmax.uu.se/naiss2023-23-3/workshops/workshop-scrnaseq"
path_file <- "data/covid/results/bioc_covid_qc_dr_int_cl.rds"
if (!dir.exists(dirname(path_file))) dir.create(dirname(path_file), recursive = TRUE)
if (fetch_data && !file.exists(path_file)) download.file(url = file.path(path_data, "covid/results/bioc_covid_qc_dr_int_cl.rds"), destfile = path_file)
alldata <- readRDS(path_file)
```

{{< meta ct_read_2 >}}

```{r}
ctrl.sce <- alldata[, alldata@colData$sample == "ctrl.13"]

# remove all old dimensionality reductions as they will mess up the analysis further down
reducedDims(ctrl.sce) <- NULL
```

## {{< meta ct_ref >}}

{{< meta ct_ref_1 >}}

```{r}
reference <- scPred::pbmc_1
reference
```

Convert to a SCE object.

```{r}
ref.sce <- Seurat::as.SingleCellExperiment(reference)
```

{{< meta ct_ref_2 >}}

```{r}
# Normalize
ref.sce <- computeSumFactors(ref.sce)
ref.sce <- logNormCounts(ref.sce)

# Variable genes
var.out <- modelGeneVar(ref.sce, method = "loess")
hvg.ref <- getTopHVGs(var.out, n = 1000)

# Dim reduction
ref.sce <- runPCA(ref.sce,
    exprs_values = "logcounts", scale = T,
    ncomponents = 30, subset_row = hvg.ref
)
ref.sce <- runUMAP(ref.sce, dimred = "PCA")
```

```{r}
#| fig-height: 5
#| fig-width: 6

plotReducedDim(ref.sce, dimred = "UMAP", colour_by = "cell_type")
```

{{< meta ct_ref_3 >}}

```{r}
# Normalize
ctrl.sce <- computeSumFactors(ctrl.sce)
ctrl.sce <- logNormCounts(ctrl.sce)

# Variable genes
var.out <- modelGeneVar(ctrl.sce, method = "loess")
hvg.ctrl <- getTopHVGs(var.out, n = 1000)

# Dim reduction
ctrl.sce <- runPCA(ctrl.sce, exprs_values = "logcounts", scale = T, ncomponents = 30, subset_row = hvg.ctrl)
ctrl.sce <- runUMAP(ctrl.sce, dimred = "PCA")
```

```{r}
#| fig-height: 5
#| fig-width: 6

plotReducedDim(ctrl.sce, dimred = "UMAP", colour_by = "louvain_SNNk15")
```

## scMap

The scMap package is one method for projecting cells from a scRNA-seq experiment on to the cell-types or individual cells identified in a different experiment. It can be run on different levels, either projecting by cluster or by single cell, here we will try out both.

For scmap cell type labels must be stored in the `cell_type1` column of the `colData` slots, and gene ids that are consistent across both datasets must be stored in the `feature_symbol` column of the `rowData` slots.

### scMap cluster

```{r}
# add in slot cell_type1
ref.sce@colData$cell_type1 <- ref.sce@colData$cell_type
# create a rowData slot with feature_symbol
rd <- data.frame(feature_symbol = rownames(ref.sce))
rownames(rd) <- rownames(ref.sce)
rowData(ref.sce) <- rd

# same for the ctrl dataset
# create a rowData slot with feature_symbol
rd <- data.frame(feature_symbol = rownames(ctrl.sce))
rownames(rd) <- rownames(ctrl.sce)
rowData(ctrl.sce) <- rd
```

Then we can select variable features in both datasets.

```{r}
# select features
counts(ctrl.sce) <- as.matrix(counts(ctrl.sce))
logcounts(ctrl.sce) <- as.matrix(logcounts(ctrl.sce))
ctrl.sce <- selectFeatures(ctrl.sce, suppress_plot = TRUE)

counts(ref.sce) <- as.matrix(counts(ref.sce))
logcounts(ref.sce) <- as.matrix(logcounts(ref.sce))
ref.sce <- selectFeatures(ref.sce, suppress_plot = TRUE)
```

Then we need to index the reference dataset by cluster, default is the clusters in `cell_type1`.

```{r}
ref.sce <- indexCluster(ref.sce)
```

Now we project the Covid-19 dataset onto that index.

```{r}
project_cluster <- scmapCluster(
    projection = ctrl.sce,
    index_list = list(
        ref = metadata(ref.sce)$scmap_cluster_index
    )
)

# projected labels
table(project_cluster$scmap_cluster_labs)
```

Then add the predictions to metadata and plot UMAP.

```{r}
#| fig-height: 5
#| fig-width: 6

# add in predictions
ctrl.sce@colData$scmap_cluster <- project_cluster$scmap_cluster_labs

plotReducedDim(ctrl.sce, dimred = "UMAP", colour_by = "scmap_cluster")
```

## scMap cell

We can instead index the refernce data based on each single cell and project our data onto the closest neighbor in that dataset.

```{r}
ref.sce <- indexCell(ref.sce)
```

Again we need to index the reference dataset.

```{r}
project_cell <- scmapCell(
    projection = ctrl.sce,
    index_list = list(
        ref = metadata(ref.sce)$scmap_cell_index
    )
)
```

We now get a table with index for the 5 nearest neigbors in the reference dataset for each cell in our dataset. We will select the celltype of the closest neighbor and assign it to the data.

```{r}
cell_type_pred <- colData(ref.sce)$cell_type1[project_cell$ref[[1]][1, ]]
table(cell_type_pred)
```

Then add the predictions to metadata and plot umap.

```{r}
#| fig-height: 5
#| fig-width: 6

# add in predictions
ctrl.sce@colData$scmap_cell <- cell_type_pred

plotReducedDim(ctrl.sce, dimred = "UMAP", colour_by = "scmap_cell")
```

Plot both:

```{r}
#| fig-height: 4
#| fig-width: 10

wrap_plots(
    plotReducedDim(ctrl.sce, dimred = "UMAP", colour_by = "scmap_cluster"),
    plotReducedDim(ctrl.sce, dimred = "UMAP", colour_by = "scmap_cell"),
    ncol = 2
)
```

## {{< meta ct_scpred >}}

{{< meta ct_scpred_1 >}}

scPred works with Seurat objects, so we will convert both objects to seurat objects. You may see a lot of warnings about renaming things, but as long as you do not see an Error, you should be fine.

```{r}
suppressPackageStartupMessages(library(Seurat))

reference <- Seurat::as.Seurat(ref.sce)
ctrl <- Seurat::as.Seurat(ctrl.sce)
```

The loadings matrix is lost when converted to Seurat object, and scPred needs that information. So we need to rerun PCA with Seurat and the same hvgs.

```{r}
VariableFeatures(reference) <- hvg.ref
reference <- reference %>%
    ScaleData(verbose = F) %>%
    RunPCA(verbose = F)

VariableFeatures(ctrl) <- hvg.ctrl
ctrl <- ctrl %>%
    ScaleData(verbose = F) %>%
    RunPCA(verbose = F)
```

```{r}
reference <- getFeatureSpace(reference, "cell_type")
reference <- trainModel(reference)
```

{{< meta ct_scpred_1 >}}

```{r}
get_scpred(reference)
```

{{< meta ct_scpred_3 >}}

```{r}
ctrl <- scPredict(ctrl, reference)
```

```{r}
#| fig-height: 5
#| fig-width: 7

DimPlot(ctrl, group.by = "scpred_prediction", label = T, repel = T) + NoAxes()
```

{{< meta ct_scpred_4 >}}

```{r}
#| fig-height: 5
#| fig-width: 7

ggplot(ctrl@meta.data, aes(x = louvain_SNNk15, fill = scpred_prediction)) +
    geom_bar() +
    theme_classic()
```

Add the predictions into the SCE object

```{r}
ctrl.sce@colData$scpred_prediction <- ctrl$scpred_prediction
```

## {{< meta ct_compare >}}

{{< meta ct_compare_1 >}}

```{r}
crossTab(ctrl, "scmap_cell", "scpred_prediction")
```

## {{< meta ct_gsea >}}

{{< meta ct_gsea_1 >}}

### {{< meta ct_gsea_deg >}}

{{< meta ct_gsea_deg_1 >}}

```{r}
# run differential expression in our dataset, using clustering at resolution 0.3
DGE_list <- scran::findMarkers(
    x = alldata,
    groups = as.character(alldata@colData$louvain_SNNk15),
    pval.type = "all",
    min.prop = 0
)
```

```{r}
# Compute differential gene expression in reference dataset (that has cell annotation)
ref_DGE <- scran::findMarkers(
    x = ref.sce,
    groups = as.character(ref.sce@colData$cell_type),
    pval.type = "all",
    direction = "up"
)

# Identify the top cell marker genes in reference dataset
# select top 50 with hihgest foldchange among top 100 signifcant genes.
ref_list <- lapply(ref_DGE, function(x) {
    x$logFC <- rowSums(as.matrix(x[, grep("logFC", colnames(x))]))
    x %>%
        as.data.frame() %>%
        filter(p.value < 0.01) %>%
        top_n(-100, p.value) %>%
        top_n(50, logFC) %>%
        rownames()
})

unlist(lapply(ref_list, length))
```


{{< meta ct_gsea_deg_2 >}}

```{r}
suppressPackageStartupMessages(library(fgsea))

# run fgsea for each of the clusters in the list
res <- lapply(DGE_list, function(x) {
    x$logFC <- rowSums(as.matrix(x[, grep("logFC", colnames(x))]))
    gene_rank <- setNames(x$logFC, rownames(x))
    fgseaRes <- fgsea(pathways = ref_list, stats = gene_rank, nperm = 10000)
    return(fgseaRes)
})
names(res) <- names(DGE_list)

# You can filter and resort the table based on ES, NES or pvalue
res <- lapply(res, function(x) {
    x[x$pval < 0.1, ]
})
res <- lapply(res, function(x) {
    x[x$size > 2, ]
})
res <- lapply(res, function(x) {
    x[order(x$NES, decreasing = T), ]
})
res
```

{{< meta ct_gsea_deg_3 >}}

```{r}
#| fig-height: 4
#| fig-width: 10

new.cluster.ids <- unlist(lapply(res, function(x) {
    as.data.frame(x)[1, 1]
}))

alldata@colData$ref_gsea <- new.cluster.ids[as.character(alldata@colData$louvain_SNNk15)]

wrap_plots(
    plotReducedDim(alldata, dimred = "UMAP", colour_by = "louvain_SNNk15"),
    plotReducedDim(alldata, dimred = "UMAP", colour_by = "ref_gsea"),
    ncol = 2
)
```

{{< meta ct_gsea_deg_4 >}}

```{r}
#| fig-height: 3.5
#| fig-width: 12

ctrl.sce@colData$ref_gsea <- alldata@colData$ref_gsea[alldata@colData$sample == "ctrl.13"]

wrap_plots(
    plotReducedDim(ctrl.sce, dimred = "UMAP", colour_by = "ref_gsea"),
    plotReducedDim(ctrl.sce, dimred = "UMAP", colour_by = "scmap_cell"),
    plotReducedDim(ctrl.sce, dimred = "UMAP", colour_by = "scpred_prediction"),
    ncol = 3
)
```

### {{< meta ct_gsea_annot >}}

{{< meta ct_gsea_annot_1 >}}

```{r}
path_file <- file.path("data/human_cell_markers.txt")
if (!file.exists(path_file)) download.file(file.path(path_data, "human_cell_markers.txt"), destfile = path_file)
```

```{r}
markers <- read.delim("data/human_cell_markers.txt")
markers <- markers[markers$speciesType == "Human", ]
markers <- markers[markers$cancerType == "Normal", ]

# Filter by tissue (to reduce computational time and have tissue-specific classification)
# sort(unique(markers$tissueType))
# grep("blood",unique(markers$tissueType),value = T)
# markers <- markers [ markers$tissueType %in% c("Blood","Venous blood",
#                                                "Serum","Plasma",
#                                                "Spleen","Bone marrow","Lymph node"), ]


# remove strange characters etc.
celltype_list <- lapply(unique(markers$cellName), function(x) {
    x <- paste(markers$geneSymbol[markers$cellName == x], sep = ",")
    x <- gsub("[[]|[]]| |-", ",", x)
    x <- unlist(strsplit(x, split = ","))
    x <- unique(x[!x %in% c("", "NA", "family")])
    x <- casefold(x, upper = T)
})
names(celltype_list) <- unique(markers$cellName)
# celltype_list <- lapply(celltype_list , function(x) {x[1:min(length(x),50)]} )
celltype_list <- celltype_list[unlist(lapply(celltype_list, length)) < 100]
celltype_list <- celltype_list[unlist(lapply(celltype_list, length)) > 5]
```

```{r}
# run fgsea for each of the clusters in the list
res <- lapply(DGE_list, function(x) {
    x$logFC <- rowSums(as.matrix(x[, grep("logFC", colnames(x))]))
    gene_rank <- setNames(x$logFC, rownames(x))
    fgseaRes <- fgsea(pathways = celltype_list, stats = gene_rank, nperm = 10000)
    return(fgseaRes)
})
names(res) <- names(DGE_list)

# You can filter and resort the table based on ES, NES or pvalue
res <- lapply(res, function(x) {
    x[x$pval < 0.01, ]
})
res <- lapply(res, function(x) {
    x[x$size > 5, ]
})
res <- lapply(res, function(x) {
    x[order(x$NES, decreasing = T), ]
})

# show top 3 for each cluster.
lapply(res, head, 3)
```

#CT_GSEA8:

```{r}
#| fig-height: 4
#| fig-width: 10

new.cluster.ids <- unlist(lapply(res, function(x) {
    as.data.frame(x)[1, 1]
}))
alldata@colData$cellmarker_gsea <- new.cluster.ids[as.character(alldata@colData$louvain_SNNk15)]

wrap_plots(
    plotReducedDim(alldata, dimred = "UMAP", colour_by = "cellmarker_gsea"),
    plotReducedDim(alldata, dimred = "UMAP", colour_by = "ref_gsea"),
    ncol = 2
)
```

:::{.callout-note title="Discuss"}
{{< meta ct_gsea_annot_2 >}}
:::

{{< meta ct_gsea_annot_3 >}}

{{< meta ct_save >}}

```{r}
saveRDS(ctrl.sce, "data/covid/results/bioc_covid_qc_dr_int_cl_ct-ctrl13.rds")
```

## {{< meta session >}}

<details>
  <summary>Click here</summary>

```{r}
sessionInfo()
```

</details>
